---
title: "SpacTempGithub"
output: html_document
date: "2026-05-13"
---
```{r}
getwd()
```

pipelines are as follows:
- for phenological (seasonal sampling) + diel sampling bioinformatic pipeline - file is titled "sm.di.pipeline.Rmd
- for IESS bioinformatic pipeline - file is titled "ie.pipeline.Rmd" 

uploading ps.cleaner (raw files) 
iess.ps.cleaner = ps.cleaner for IESS samples (internal/external spatial survey)
ps.ps.cleaner = ps.cleaner for PS samples (phenological)
di.ps.cleaner = ps.cleaner for DI samples (diel)

raw counts..need to clr tranform them
```{r}
iess.ps.cleaner <- readRDS("CH1_IE_16S_ps.cleaner.RDS")
ps.ps.cleaner <- readRDS("sm1.ps.cleaner.rds")
di.ps.cleaner <- readRDS("diel.ps.cleaner.rds")
```

ensuring raw counts
```{r}
otu_table(iess.ps.cleaner)[1:5, 1:5]
otu_table(ps.ps.cleaner)[1:5, 1:5]
otu_table(di.ps.cleaner)[1:5, 1:5]
```

clr transformation for each 
```{r}
iess.ps.cleaner.clr <- microbiome::transform(iess.ps.cleaner,'clr')
iess.seqtab.clr <- data.frame(otu_table(iess.ps.cleaner.clr))

ps.ps.cleaner.clr <- microbiome::transform(ps.ps.cleaner,'clr')
ps.seqtab.clr <- data.frame(otu_table(ps.ps.cleaner.clr))

di.ps.cleaner.clr <- microbiome::transform(di.ps.cleaner,'clr')
di.seqtab.clr <- data.frame(otu_table(di.ps.cleaner.clr))
```

ensuring clr tranformation and saving dataframes
```{r}
otu_table(iess.ps.cleaner.clr)[1:5, 1:5]
otu_table(ps.ps.cleaner.clr)[1:5, 1:5]
otu_table(di.ps.cleaner.clr)[1:5, 1:5]
```
good^
```{r}
saveRDS(iess.ps.cleaner.clr, "iess.ps.cleaner.clr.rds")
saveRDS(ps.ps.cleaner.clr, "ps.ps.cleaner.clr.rds")
saveRDS(di.ps.cleaner.clr, "di.ps.cleaner.clr.rds")
```
all saved & uploaded into github to start back here! 

Now let's start with the stats
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~SURVEY STATS & Figures~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
Note: Figure 2 Map (not made in r, made on arcmap gis)

PCA 1 - GROUP BY MAT
```{r}
pca.df <- read.csv("pca.df.update.csv")
```

```{r}
# create factor for grouping
pca.df$mat_group <- factor(pca.df$mat, labels = c("No mats", "Mats"))
```

```{r}
# select only numeric environmental variables
env_vars <- pca.df %>%
  select(cond, sal, TDS, pH, RDO.Sat, temp, DOC)
```

```{r}
# run PCA (scale is important)
pca <- prcomp(env_vars, scale. = TRUE)
```

```{r}
biplot(pca)
```
Figure 3A - labels manually added later 
```{r}
library(factoextra)
pca_mat1.2 <- fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage = pca.df$mat_group,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  label = "none",   # <-- removes labels, keeps arrows
  col.var = "black",
  pointsize = 3,
  alpha.ind = 0.8,
  repel = TRUE,
  legend.title = "Mat Presence",
  title = "PCA of Water Quality Parameters",
  select.var = list(name = c("pH", "cond", "sal", "DOC", "temp", "RDO.Sat"))
) +
  theme_minimal(base_size = 14)

pca_mat1.2

```
Notes on PCA1: 

- Dim1 = 57.7% of the variation
- Dim2 = 22.2% of the variation
- together, almost 80% 

- Mats (teal triangles) cluster more toward the right side of Dim1
- No mats (red circles) are more spread out and include more of the left side
- mat presence is associated with a particular water chemistry signature, not random conditions

- Variables pointing to the RIGHT (positive Dim1): cond (conductivity), TDS, sal (salinity), pH (slightly up-right)

- These are tightly grouped  ~  positively correlated with each other
- So Dim1 is salinity / ionic strength gradient.

-Sites with mats tend to occur where: higher conductivity, higher TDS, higher salinity, slightly higher pH

- Variables pointing DOWN (negative Dim2): temp, RDO.Sat (dissolved oxygen saturation)
- So Dim2 is mostly a temperature / oxygen gradient.
- Lower Dim2 values = warmer water, lower DO saturation

- DOC arrow points left
- DOC is opposite of salinity/TDS/cond
- So:High DOC = low salinity & conductivity, Low DOC = higher salinity waters
- makes sense!

```{r}
#save as PDF
ggsave(
  filename = "pca.mat.nolabels.pdf",
  plot = pca_mat1.2,
  width = 8,
  height = 6
)
```

statistical test to back this up: PERMANOVA
```{r}
env_vars <- pca.df[, c("cond", "TDS", "sal", "pH", "DOC", "temp", "RDO.Sat")]
library(vegan)
env_scaled <- scale(env_vars)
env_dist <- dist(env_scaled, method = "euclidean")
adonis2(env_dist ~ mat_group, 
        data = pca.df,
        permutations = 9999)
```
not stat supported by a PERMANOVA boo hoo
beta dispersion assumption passes ~
```{r}
#BETADISPERSIONCHECK
disp1 <- betadisper(env_dist, pca.df$mat_group)
anova(disp1)
```
PCA 2 - GROUP BY MONTH!

```{r}
# create factor for grouping
pca.df$month_group <- factor(
  pca.df$month,
  levels = c("Mar_2023", "June_2023", "July_2023"),
  labels = c("March", "June", "July")
)

# select only numeric environmental variables
env_vars <- pca.df %>%
  select(cond, sal, TDS, pH, RDO.Sat, temp, DOC)

# run PCA (scale is important)
pca <- prcomp(env_vars, scale. = TRUE)
```

```{r}
biplot(pca)
```
Figure 3B ~ labels manually fixed later
```{r}
pca_month <- fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage = pca.df$month_group,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  label = "var",   
  col.var = "black",
  pointsize = 3,
  alpha.ind = 0.8,
  repel = TRUE,
  legend.title = "Month",
  title = "PCA of Water Quality Parameters",
  select.var = list(name = c("pH", "cond", "TDS", "sal", "DOC", "temp", "RDO.Sat"))
) +
  theme_minimal(base_size = 14)

pca_month

```
```{r}
# Optionally, save as PDF
ggsave(
  filename = "pca.month.nolabel.pdf",
  plot = pca_month,
  width = 8,
  height = 6
)
```


statistical test to back this up: PERMANOVA
```{r}
env_vars <- pca.df[, c("cond", "TDS", "sal", "pH", "DOC", "temp", "RDO.Sat")]
library(vegan)
env_scaled <- scale(env_vars)
env_dist <- dist(env_scaled, method = "euclidean")
adonis2(env_dist ~ month_group, 
        data = pca.df,
        permutations = 9999)

#BETADISPERSIONCHECK
disp2 <- betadisper(env_dist, pca.df$month_group)
anova(disp2)
#passes beta dispersion result!

#pariwise test
# install if needed
# install.packages("devtools")
# devtools::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")

library(pairwiseAdonis)

# Run pairwise PERMANOVA
pairwise.adonis2(env_dist ~ month_group, data = pca.df, permutations = 9999, p.adjust.m = "BH")

```
differences seen between 3 sampling months!

PCA 1 - GROUP BY WATER TYPE
```{r}
# create factor for grouping
pca.df$water_group <- factor(
  pca.df$water,
  levels = c("sound", "lagoon", "channel", "river estuary"),
  labels = c("S", "L", "C", "RE")
)
```

Figure 3C ~ labels manually added later
```{r}
pca_water <- fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage = pca.df$water_group,
  addEllipses = FALSE,
  #ellipse.level = 0.95,
  label = "none",   
  col.var = "black",
  pointsize = 3,
  alpha.ind = 0.8,
  repel = TRUE,
  legend.title = "Water Type",
  title = "",
  select.var = list(name = c("pH", "cond", "TDS", "sal", "DOC", "temp", "RDO.Sat"))
) +
  theme_minimal(base_size = 14)

pca_water
```

To save
```{r}
# Optionally, save as PDF
ggsave(
  filename = "pca.water.nolab.noellipses.pdf",
  plot = pca_water,
  width = 8,
  height = 6
)
```

```{r}
adonis2(env_dist ~ water_group, 
        data = pca.df,
        permutations = 9999)

#BETADISPERSIONCHECK
disp5 <- betadisper(env_dist, pca.df$water_group)
anova(disp5)

# Run pairwise PERMANOVA
#pairwise.adonis2(env_dist ~ month_group, data = pca.df, permutations = 9999, p.adjust.m = "BH")
```
*not stat sig different based on water type~ beta disp passes

GROUP BY REGION (NEARBY/WATER CLUMPING SPOTS)
```{r}
library(dplyr)

# create factor for geographic grouping
pca.df <- pca.df %>%
  mutate(site_group = case_when(
    site %in% c("HH") ~ "Western_Sound",
    site %in% c("JI") ~ "Eastern_Sound",
    site %in% c("MM", "SBE", "SBC", "SB") ~ "Central_Lagoon_Sound",
    site %in% c("TM", "BS", "FM") ~ "Northern_Lagoon_Sound_Channel",
    site %in% c("PI", "TI") ~ "Northern_River_Estuary",
    TRUE ~ "Other"
  ))

pca.df$site_group <- factor(pca.df$site_group,
                            levels = c("Western_Sound",
                                       "Eastern_Sound",
                                       "Central_Lagoon_Sound",
                                       "Northern_Lagoon_Sound_Channel",
                                       "Northern_River_Estuary"),
                            labels = c("WS",
                                       "ES",
                                       "CLS",
                                       "NLSC",
                                       "NRE"))


```
 
Figure 3D ~ labels manually added later
```{r}
pca_loc <- fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage = pca.df$site_group,
  addEllipses = FALSE,
  label = "none",   
  col.var = "black",
  pointsize = 3,
  alpha.ind = 0.8,
  repel = TRUE,
  legend.title = "region",
  title = "",
  select.var = list(name = c("pH", "cond", "TDS", "sal", "DOC", "temp", "RDO.Sat"))
) +
  theme_minimal(base_size = 14)

pca_loc
```
To save
```{r}
# Optionally, save as PDF
ggsave(
  filename = "pca.loc.nolab.noellipses.pdf",
  plot = pca_loc,
  width = 8,
  height = 6
)
```
 
stats check 
```{r}
adonis2(env_dist ~ site_group, 
        data = pca.df,
        permutations = 9999)

#BETADISPERSIONCHECK
disp6 <- betadisper(env_dist, pca.df$site_group)
anova(disp6)

# Run pairwise PERMANOVA
#pairwise.adonis2(env_dist ~ month_group, data = pca.df, permutations = 9999, p.adjust.m = "BH")
```
*not stat sig different based on region ~ beta disp passes

putting plot together ~ figure 3 final!



```{r}
library(cowplot)
pca.top <-cowplot::plot_grid(
   pca_mat1.2,
   pca_month,
  nrow = 1,
  align = "h",
  axis = "tb")
pca.top
```

```{r}
pca.bottom <- cowplot::plot_grid(
   pca_water,
   pca_loc,
  nrow = 1,
  align = "h",
  axis = "tb")
pca.bottom
```
```{r}
library(cowplot)
final.pca <- cowplot::plot_grid(
pca.top,
pca.bottom,
  nrow = 2,
  align = "h",
  axis = "tb")

```

```{r}
final.pca
```
```{r}
ggsave(
  filename = "final.pca.pdf",
  plot = final.pca,
  width = 10,
  height = 10,
  device = cairo_pdf
)

```

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~IESS Stats!~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~








