---
title: "SpacTempGithub"
output: html_document
date: "2026-05-13"
---

loading libraries
```{r}
library(stringr)
library(dplyr)
library(stats)
library(ggplot2)
library(ggpubr)
library(vegan)
library(cowplot)
library(tidyverse)
library(phyloseq)
library(microbiome)
library("funfuns")
library(pairwiseAdonis)
library(viridis)
library("microViz")
```

```{r}
getwd()
```

pipelines are as follows:
- for phenological (seasonal sampling) + diel sampling bioinformatic pipeline - file is titled "sm.di.pipeline.Rmd
- for IESS bioinformatic pipeline - file is titled "ie.pipeline.Rmd" 

uploading ps.cleaner (raw files) 
iess.ps.cleaner = ps.cleaner for IESS samples (internal/external spatial survey)
ps.ps.cleaner = ps.cleaner for PS samples (phenological)
di.ps.cleaner = ps.cleaner for DI samples (diel)

raw counts..need to clr tranform them
```{r}
iess.ps.cleaner <- readRDS("CH1_IE_16S_ps.cleaner.RDS")
ps.ps.cleaner <- readRDS("sm1.ps.cleaner.rds")
di.ps.cleaner <- readRDS("diel.ps.cleaner.rds")
```

ensuring raw counts
```{r}
otu_table(iess.ps.cleaner)[1:5, 1:5]
otu_table(ps.ps.cleaner)[1:5, 1:5]
otu_table(di.ps.cleaner)[1:5, 1:5]
```

clr transformation for each 
```{r}
iess.ps.cleaner.clr <- microbiome::transform(iess.ps.cleaner,'clr')
iess.seqtab.clr <- data.frame(otu_table(iess.ps.cleaner.clr))

ps.ps.cleaner.clr <- microbiome::transform(ps.ps.cleaner,'clr')
ps.seqtab.clr <- data.frame(otu_table(ps.ps.cleaner.clr))

di.ps.cleaner.clr <- microbiome::transform(di.ps.cleaner,'clr')
di.seqtab.clr <- data.frame(otu_table(di.ps.cleaner.clr))
```

ensuring clr tranformation and saving dataframes
```{r}
otu_table(iess.ps.cleaner.clr)[1:5, 1:5]
otu_table(ps.ps.cleaner.clr)[1:5, 1:5]
otu_table(di.ps.cleaner.clr)[1:5, 1:5]
```
good^
```{r}
saveRDS(iess.ps.cleaner.clr, "iess.ps.cleaner.clr.rds")
saveRDS(ps.ps.cleaner.clr, "ps.ps.cleaner.clr.rds")
saveRDS(di.ps.cleaner.clr, "di.ps.cleaner.clr.rds")
```
all saved & uploaded into github to start back here! 

Now let's start with the stats
~~~~~~~~~~~~~~~SURVEY STATS & Figures~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
Note: Figure 2 Map (not made in r, made on arcmap gis)

PCA 1 - GROUP BY MAT
```{r}
pca.df <- read.csv("pca.df.update.csv")
```

```{r}
# create factor for grouping
pca.df$mat_group <- factor(pca.df$mat, labels = c("No mats", "Mats"))
```

```{r}
# select only numeric environmental variables
env_vars <- pca.df %>%
  select(cond, sal, TDS, pH, RDO.Sat, temp, DOC)
```

```{r}
# run PCA (scale is important)
pca <- prcomp(env_vars, scale. = TRUE)
```

```{r}
biplot(pca)
```
Figure 3A - labels manually added later 
```{r}
library(factoextra)
pca_mat1.2 <- fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage = pca.df$mat_group,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  label = "none",   # <-- removes labels, keeps arrows
  col.var = "black",
  pointsize = 3,
  alpha.ind = 0.8,
  repel = TRUE,
  legend.title = "Mat Presence",
  title = "PCA of Water Quality Parameters",
  select.var = list(name = c("pH", "cond", "sal", "DOC", "temp", "RDO.Sat"))
) +
  theme_minimal(base_size = 14)

pca_mat1.2

```
Notes on PCA1: 

- Dim1 = 57.7% of the variation
- Dim2 = 22.2% of the variation
- together, almost 80% 

- Mats (teal triangles) cluster more toward the right side of Dim1
- No mats (red circles) are more spread out and include more of the left side
- mat presence is associated with a particular water chemistry signature, not random conditions

- Variables pointing to the RIGHT (positive Dim1): cond (conductivity), TDS, sal (salinity), pH (slightly up-right)

- These are tightly grouped  ~  positively correlated with each other
- So Dim1 is salinity / ionic strength gradient.

-Sites with mats tend to occur where: higher conductivity, higher TDS, higher salinity, slightly higher pH

- Variables pointing DOWN (negative Dim2): temp, RDO.Sat (dissolved oxygen saturation)
- So Dim2 is mostly a temperature / oxygen gradient.
- Lower Dim2 values = warmer water, lower DO saturation

- DOC arrow points left
- DOC is opposite of salinity/TDS/cond
- So:High DOC = low salinity & conductivity, Low DOC = higher salinity waters
- makes sense!

```{r}
#save as PDF
ggsave(
  filename = "pca.mat.nolabels.pdf",
  plot = pca_mat1.2,
  width = 8,
  height = 6
)
```

statistical test to back this up: PERMANOVA
```{r}
env_vars <- pca.df[, c("cond", "TDS", "sal", "pH", "DOC", "temp", "RDO.Sat")]
library(vegan)
env_scaled <- scale(env_vars)
env_dist <- dist(env_scaled, method = "euclidean")
adonis2(env_dist ~ mat_group, 
        data = pca.df,
        permutations = 9999)
```
not stat supported by a PERMANOVA boo hoo
beta dispersion assumption passes ~
```{r}
#BETADISPERSIONCHECK
disp1 <- betadisper(env_dist, pca.df$mat_group)
anova(disp1)
```
PCA 2 - GROUP BY MONTH!

```{r}
# create factor for grouping
pca.df$month_group <- factor(
  pca.df$month,
  levels = c("Mar_2023", "June_2023", "July_2023"),
  labels = c("March", "June", "July")
)

# select only numeric environmental variables
env_vars <- pca.df %>%
  select(cond, sal, TDS, pH, RDO.Sat, temp, DOC)

# run PCA (scale is important)
pca <- prcomp(env_vars, scale. = TRUE)
```

```{r}
biplot(pca)
```
Figure 3B ~ labels manually fixed later
```{r}
pca_month <- fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage = pca.df$month_group,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  label = "var",   
  col.var = "black",
  pointsize = 3,
  alpha.ind = 0.8,
  repel = TRUE,
  legend.title = "Month",
  title = "PCA of Water Quality Parameters",
  select.var = list(name = c("pH", "cond", "TDS", "sal", "DOC", "temp", "RDO.Sat"))
) +
  theme_minimal(base_size = 14)

pca_month

```
```{r}
# Optionally, save as PDF
ggsave(
  filename = "pca.month.nolabel.pdf",
  plot = pca_month,
  width = 8,
  height = 6
)
```


statistical test to back this up: PERMANOVA
```{r}
env_vars <- pca.df[, c("cond", "TDS", "sal", "pH", "DOC", "temp", "RDO.Sat")]
library(vegan)
env_scaled <- scale(env_vars)
env_dist <- dist(env_scaled, method = "euclidean")
adonis2(env_dist ~ month_group, 
        data = pca.df,
        permutations = 9999)

#BETADISPERSIONCHECK
disp2 <- betadisper(env_dist, pca.df$month_group)
anova(disp2)
#passes beta dispersion result!

#pariwise test
# install if needed
# install.packages("devtools")
# devtools::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")

library(pairwiseAdonis)

# Run pairwise PERMANOVA
pairwise.adonis2(env_dist ~ month_group, data = pca.df, permutations = 9999, p.adjust.m = "BH")

```
differences seen between 3 sampling months!

PCA 1 - GROUP BY WATER TYPE
```{r}
# create factor for grouping
pca.df$water_group <- factor(
  pca.df$water,
  levels = c("sound", "lagoon", "channel", "river estuary"),
  labels = c("S", "L", "C", "RE")
)
```

Figure 3C ~ labels manually added later
```{r}
pca_water <- fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage = pca.df$water_group,
  addEllipses = FALSE,
  #ellipse.level = 0.95,
  label = "none",   
  col.var = "black",
  pointsize = 3,
  alpha.ind = 0.8,
  repel = TRUE,
  legend.title = "Water Type",
  title = "",
  select.var = list(name = c("pH", "cond", "TDS", "sal", "DOC", "temp", "RDO.Sat"))
) +
  theme_minimal(base_size = 14)

pca_water
```

To save
```{r}
# Optionally, save as PDF
ggsave(
  filename = "pca.water.nolab.noellipses.pdf",
  plot = pca_water,
  width = 8,
  height = 6
)
```

```{r}
adonis2(env_dist ~ water_group, 
        data = pca.df,
        permutations = 9999)

#BETADISPERSIONCHECK
disp5 <- betadisper(env_dist, pca.df$water_group)
anova(disp5)

# Run pairwise PERMANOVA
#pairwise.adonis2(env_dist ~ month_group, data = pca.df, permutations = 9999, p.adjust.m = "BH")
```
*not stat sig different based on water type~ beta disp passes

GROUP BY REGION (NEARBY/WATER CLUMPING SPOTS)
```{r}
library(dplyr)

# create factor for geographic grouping
pca.df <- pca.df %>%
  mutate(site_group = case_when(
    site %in% c("HH") ~ "Western_Sound",
    site %in% c("JI") ~ "Eastern_Sound",
    site %in% c("MM", "SBE", "SBC", "SB") ~ "Central_Lagoon_Sound",
    site %in% c("TM", "BS", "FM") ~ "Northern_Lagoon_Sound_Channel",
    site %in% c("PI", "TI") ~ "Northern_River_Estuary",
    TRUE ~ "Other"
  ))

pca.df$site_group <- factor(pca.df$site_group,
                            levels = c("Western_Sound",
                                       "Eastern_Sound",
                                       "Central_Lagoon_Sound",
                                       "Northern_Lagoon_Sound_Channel",
                                       "Northern_River_Estuary"),
                            labels = c("WS",
                                       "ES",
                                       "CLS",
                                       "NLSC",
                                       "NRE"))


```
 
Figure 3D ~ labels manually added later
```{r}
pca_loc <- fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage = pca.df$site_group,
  addEllipses = FALSE,
  label = "none",   
  col.var = "black",
  pointsize = 3,
  alpha.ind = 0.8,
  repel = TRUE,
  legend.title = "region",
  title = "",
  select.var = list(name = c("pH", "cond", "TDS", "sal", "DOC", "temp", "RDO.Sat"))
) +
  theme_minimal(base_size = 14)

pca_loc
```
To save
```{r}
# Optionally, save as PDF
ggsave(
  filename = "pca.loc.nolab.noellipses.pdf",
  plot = pca_loc,
  width = 8,
  height = 6
)
```
 
stats check 
```{r}
adonis2(env_dist ~ site_group, 
        data = pca.df,
        permutations = 9999)

#BETADISPERSIONCHECK
disp6 <- betadisper(env_dist, pca.df$site_group)
anova(disp6)

# Run pairwise PERMANOVA
#pairwise.adonis2(env_dist ~ month_group, data = pca.df, permutations = 9999, p.adjust.m = "BH")
```
*not stat sig different based on region ~ beta disp passes

putting plot together ~ figure 3 final!



```{r}
library(cowplot)
pca.top <-cowplot::plot_grid(
   pca_mat1.2,
   pca_month,
  nrow = 1,
  align = "h",
  axis = "tb")
pca.top
```

```{r}
pca.bottom <- cowplot::plot_grid(
   pca_water,
   pca_loc,
  nrow = 1,
  align = "h",
  axis = "tb")
pca.bottom
```
```{r}
library(cowplot)
final.pca <- cowplot::plot_grid(
pca.top,
pca.bottom,
  nrow = 2,
  align = "h",
  axis = "tb")

```

```{r}
final.pca
```
```{r}
ggsave(
  filename = "final.pca.pdf",
  plot = final.pca,
  width = 10,
  height = 10,
  device = cairo_pdf
)

```


~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~IESS Stats!~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```{r}
iess.seqtab <- data.frame(otu_table(iess.ps.cleaner))
iess.dist.all <- vegdist(iess.seqtab, method = "euclidean")
```

extract metadata
```{r}
iess.samdf <- data.frame(sample_data(iess.ps.cleaner))
```

Beta Dispersion!
```{r}
iess.bet.all1 <- betadisper(iess.dist.all,iess.samdf$Sample.Type)
anova(iess.bet.all1) 

iess.bet.all2 <- betadisper(iess.dist.all,iess.samdf$Site)
anova(iess.bet.all2) 
```
Not significant p=.1002, .9643

Beta Diversity
```{r}
iess.samdf$Sample.Type <- as.factor(iess.samdf$Sample.Type)
iess.samdf$Site <- as.factor(iess.samdf$Site)
adonis2(iess.dist.all ~ Sample.Type * Site , data=iess.samdf, permutations=9999)
```
Alpha Diversity!
alpha diversity dataframe set up: 
how did we make these? 
alpha_diversity<- data.frame(estimate_richness(ps.cleaner, split=TRUE, measures=c("Shannon","InvSimpson","Observed")))
alpha_diversity$Evenness <- alpha_diversity$Shannon / log(alpha_diversity$Observed)

^did this for each experiment and saved dataframes to be uploaded below 
ps.cleaner phyloseq object should not have transformed data (should use raw or minimally processed data for this)
*if included below then the model passes assumptions + significant results 

```{r}
iess.alpha <- data.frame(estimate_richness(iess.ps.cleaner, split=TRUE, measures=c("Shannon","InvSimpson","Observed")))
iess.alpha$Evenness <- iess.alpha$Shannon / log(iess.alpha$Observed)

sample_metadata <- data.frame(sample_data(iess.ps.cleaner))
iess.alpha <- cbind(sample_metadata, iess.alpha)

write.csv(iess.alpha,"iess.alpha.csv")
```

*next time just read in iess.alpha
model for shannon, richness, evenness, invsimpson

which model type? 

the 1|Mat accounts for samples within mats not being truly independent, especially important for iess experiment
*only included stat sig models, but ran these for each alpha div stat
*no need for interactions between variables 
```{r}
iess.alpha.2 <- lmer(InvSimpson ~ Site + Sample.Type + (1|Mat) , data = iess.alpha, REML = FALSE)
summary(iess.alpha.2)
```
```{r}
# Plot residuals against fitted values
plot(residuals(iess.alpha.2) ~ fitted(iess.alpha.2))

# Shapiro-Wilk normality test
shapiro.test(residuals(iess.alpha.2))

# Q-Q plot of residuals
qqnorm(residuals(iess.alpha.2))
qqline(residuals(iess.alpha.2))
plot_model(iess.alpha.2, type = 'diag')
#passes assumptions

```
passes model assumptions

```{r}
iess.alpha.3 <- lmer(Evenness ~ Site + Sample.Type + (1|Mat) , data = iess.alpha, REML = FALSE)
summary(iess.alpha.3)
```

```{r}
# Plot residuals against fitted values
plot(residuals(iess.alpha.3) ~ fitted(iess.alpha.3))

# Shapiro-Wilk normality test
shapiro.test(residuals(iess.alpha.3))

# Q-Q plot of residuals
qqnorm(residuals(iess.alpha.3))
qqline(residuals(iess.alpha.3))
plot_model(iess.alpha.3, type = 'diag')
#passes assumptions

```
passes model assumptions

just to ensure which one is higher 
```{r}
library(ggplot2)

ggplot(iess.alpha, aes(x = Sample.Type, y = Evenness, color = Site)) +
  geom_boxplot(position = position_dodge(width = 0.8), alpha = 0.3) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8)) +
  theme_classic() +
  labs(x = "Sample Type", y = "Evenness")


ggplot(iess.alpha, aes(x = Sample.Type, y = InvSimpson, color = Site)) +
  geom_boxplot(position = position_dodge(width = 0.8), alpha = 0.3) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8)) +
  theme_classic() +
  labs(x = "Sample Type", y = "InvSimpson")
```
*internal evenness and inv simpson  is higher
*plots above not included in paper, purely for result visualization

Diff Abund! 
(1) sample type
```{r}
ie.diff.abund1 <- ancombc2(
  data = iess.ps.cleaner,      # your phyloseq object
  tax_level = "Family",      # or "Genus", etc.
  fix_formula = "Sample.Type",  # variable to test
  p_adj_method = "holm",
  group = "Sample.Type",        # group variable for pairwise tests
  struc_zero = TRUE,         # detects true absences
  neg_lb = TRUE,
  em_control = list(
    tol = 1e-5,
    max_iter = 100,
    sensitivity = TRUE
  ),
  alpha = 0.05,
  global = TRUE
)
```
```{r}
#structural zero results (need to clean up in excel)
View(ie.diff.abund1$zero_ind)
#write.csv(ie.diff.abund1$zero_ind, "ie.diff.abund1_strzeros.csv", row.names = TRUE)
#sensitivity results (scores are low, this is why we negate global and pairwise results and focus on the robust structural zero results above)
View(ie.diff.abund1$pseudo_sens_tab)
#write.csv(ie.diff.abund1$pseudo_sens_tab, "ie.diff.abund1_sens.csv", row.names = TRUE)
```

(1) site
```{r}
ie.diff.abund2 <- ancombc2(
  data = iess.ps.cleaner,      # your phyloseq object
  tax_level = "Family",      # or "Genus", etc.
  fix_formula = "Site",  # variable to test
  p_adj_method = "holm",
  group = "Site",        # group variable for pairwise tests
  struc_zero = TRUE,         # detects true absences
  neg_lb = TRUE,
  em_control = list(
    tol = 1e-5,
    max_iter = 100,
    sensitivity = TRUE
  ),
  alpha = 0.05,
  global = TRUE
)
```
```{r}
#structural zero results (need to clean up in excel)
View(ie.diff.abund2$zero_ind)
#write.csv(ie.diff.abund2$zero_ind, "ie.diff.abund2_strzeros.csv", row.names = TRUE)
#sensitivity results (scores are low, this is why we negate global and pairwise results and focus on the robust structural zero results above)
View(ie.diff.abund2$pseudo_sens_tab)
#write.csv(ie.diff.abund2$pseudo_sens_tab, "ie.diff.abund2_sens.csv", row.names = TRUE)
```


Nutrient Stats/correlations! iess.alpha (has nutrient data) 


~look at differences of TOM, TOC, TN between sample types (internal vs external samples) &  across sites

(1) TOM

# Organic Matter
mod_OM <- lmer(X..Organic.Matter ~ Sample.Type + Site + (1|Mat),
               data = iess.alpha)
#assumptions
# Plot residuals against fitted values
plot(residuals(mod_OM) ~ fitted(mod_OM))

# Shapiro-Wilk normality test
shapiro.test(residuals(mod_OM))

# Q-Q plot of residuals
qqnorm(residuals(mod_OM))
qqline(residuals(mod_OM))
plot_model(mod_OM, type = 'diag')
#does not pass so wilcoxin paired test
`

~ Using wilcoxin test for TOM because PAIRED internal/external samples & lme model did not pass assumptions 
```{r}
wilcox.test(X..Organic.Matter ~ Sample.Type,
            paired = TRUE,
            data = iess.alpha)

wilcox.test(X..Organic.Matter ~ Site,
            paired = TRUE,
            data = iess.alpha)
```
```{r}
#not adding to paper
ggplot(iess.alpha, aes(x = Site, y = X..Organic.Matter , color = Site)) +
  geom_boxplot(position = position_dodge(width = 0.8), alpha = 0.3) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8)) +
  theme_classic() +
  labs(x = "site", y = "tom")
```



(2) TN
```{r}
#TN
mod_N <- lmer(X..Nitrogen ~ Sample.Type + Site + (1|Mat),
               data = iess.alpha)
#assumptions
# Plot residuals against fitted values
plot(residuals(mod_N) ~ fitted(mod_N))

# Shapiro-Wilk normality test
shapiro.test(residuals(mod_N))

# Q-Q plot of residuals
qqnorm(residuals(mod_N))
qqline(residuals(mod_N))
plot_model(mod_N, type = 'diag')

```


```{r}
library(lmerTest)

mod_N <- lmer(X..Nitrogen ~ Sample.Type + Site + (1|Mat),
              data = iess.alpha)

summary(mod_N)
```

(2) TOC
```{r}
#TOC
mod_TOC <- lmer(X..Organic.Carbon ~ Sample.Type + Site + (1|Mat),
               data = iess.alpha)
#assumptions
# Plot residuals against fitted values
plot(residuals(mod_TOC) ~ fitted(mod_TOC))

# Shapiro-Wilk normality test
shapiro.test(residuals(mod_TOC))

# Q-Q plot of residuals
qqnorm(residuals(mod_TOC))
qqline(residuals(mod_TOC))
plot_model(mod_TOC, type = 'diag')

```
```{r}
mod_TOC <- lmer(X..Organic.Carbon ~ Sample.Type + Site + (1|Mat),
               data = iess.alpha)
summary(mod_TOC)
```
```{r}
#not adding to paper
ggplot(iess.alpha, aes(x = Sample.Type, y = X..Organic.Carbon, color = Site)) +
  geom_boxplot(position = position_dodge(width = 0.8), alpha = 0.3) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8)) +
  theme_classic() +
  labs(x = "sample type", y = "toc")
```

~nutrient correlations 
to see families across the entire experiment:
```{r}
#note:iessmdffam ~ calculated and saved below ..need to  read back in here
iessmdffam %>%
  group_by(Family) %>%
  summarise(Total_Abundance = sum(Abundance)) %>%
  arrange(desc(Total_Abundance))
```

These are the >2% most relative abundant families (percentage) 
```{r}
top_families <- iessmdffam %>%
  group_by(Family) %>%
  summarise(Abundance = sum(Abundance)) %>%
  mutate(Percentage = (Abundance / sum(Abundance)) * 100) %>%
  filter(Percentage > 2) %>%
  arrange(desc(Percentage))

top_families
```

Calculating correlation coefficients based on Pearson
```{r}
# Load libraries
library(dplyr)
library(tidyr)
library(ggplot2)

#iess.alpha
#iessmdffam

# Merge dataframes 
merged.iess <- merge(iess.alpha, iessmdffam, by = "Sample")

#Define top families (this is taxa w > 2% relative abundance) ---
specific_family <- c("Rhodobacteraceae", "Desulfocapsaceae", "Flavobacteriaceae",
                     "Saprospiraceae","Woeseiaceae","Pirellulaceae",
                     "Halieaceae","A4b", "Cyclobacteriaceae", "Desulfosarcinaceae", "Puniceicoccaceae")

# Filter for top families & remove NAs 
merged_clean <- merged.iess %>%
  filter(Family %in% specific_family) %>%
  drop_na(Abundance, X..Organic.Matter.x, X..Organic.Carbon.x, X..Nitrogen.x)

#  Pivot nutrients to long format 
corr_df_long <- merged_clean %>%
  pivot_longer(
    cols = c(X..Organic.Matter.x, X..Organic.Carbon.x, X..Nitrogen.x),
    names_to = "Variable",
    values_to = "Nutrient"
  )

# Calculate correlations + p-values per family × nutrient
corr_family <- corr_df_long %>%
  group_by(Family, Variable) %>%
  summarize(
    Correlation_Coefficient = ifelse(
      length(unique(Nutrient)) > 1,
      cor.test(Abundance, Nutrient, method = "pearson")$estimate,
      NA
    ),
    
    P_value = ifelse(
      length(unique(Nutrient)) > 1,
      cor.test(Abundance, Nutrient, method = "pearson")$p.value,
      NA
    ),
    
    .groups = "drop"
  )
#write.csv(corr_family, "iess.corr_family.csv")

```
~sig stars added later~
corr_fam uploaded to github for easy readability of sig correlations 


HEATMAP #decided not to add to final paper!
```{r}
#make family in alphebetical order
corr_family$Family <- factor(
  corr_family$Family,
  levels = sort(unique(corr_family$Family), decreasing = TRUE)
)

iessheatmap_topfamily <- ggplot(corr_family, aes(x = Variable, y = Family, fill = Correlation_Coefficient)) +
  geom_tile(color = "white") +
  
  scale_fill_gradientn(
  colours = c(
    "#A6CEE3",  # light blue
    "#1F78B4",  # dark blue
    "#B2DF8A",  # light green
    "#33A02C"   # dark green
  ),
  name = "Pearson's Correlation"
) +
  scale_y_discrete(position = "right") +
  labs(x = "Nutrients", y = "Top Families") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    panel.border = element_blank(),
    legend.position = "right"
  )

iessheatmap_topfamily
```
*results arent super interesting (only 1 stat) so scratch above heatmap for final paper
*corr_fam uploaded to github for easy readability of sig correlations (just in case) 



Bar plots!
Need to sum groups (ie + site)
```{r}
library(microshades)
iess.ps.cleaner.rel <- transform_sample_counts(iess.ps.cleaner, function(x) x / sum(x))
fam.aggreg1 <- tax_glom(iess.ps.cleaner.rel, "Family")
iessmdffam<- prep_mdf(fam.aggreg1, subgroup_level = "Family")
```
```{r}
write.csv(iessmdffam, "iess.mdffam.csv")
```

site example below: 
*repeated for sample.type 
```{r}
library(dplyr)
iessfam_site <- iessmdffam %>%
  group_by(Site, Family) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

#convert to relative abundance within each transect
iessfam_site <- iessfam_site %>%
  group_by(Site) %>%
  mutate(Percentage = (Abundance / sum(Abundance)) * 100)

#grouping rules (same as above) applied
#families less than 1% relatively abundant grouped in "rare families" category 
#families only assigned at higher taxonomic rankings (phyla, class, order, ect) grouped into "Assigned at Higher Taxonomy" category
#note: some families were "unknown" meaning silva database was unable to find a match, so there should also be an "unknown" category unless all samples had identification to the family level

iessfam_site$Family_grouped <- ifelse(
  grepl("_", iessfam_site$Family),
  "Assigned at Higher Taxonomy",
  ifelse(iessfam_site$Percentage < 1, "Rare Families", iessfam_site$Family)
)

#re-sum after grouping (bc grouped categories are now collapsed)
iess_site_grouped <- iessfam_site %>%
  group_by(Site, Family_grouped) %>%
  summarise(Percentage = sum(Percentage), .groups = "drop")

#consistent stacking order based on overall abundance!
family_order <- iess_site_grouped %>%
  group_by(Family_grouped) %>%
  summarise(Total = sum(Percentage)) %>%
  arrange(Total) %>%
  pull(Family_grouped)

iess_site_grouped$Family_grouped <- factor(
  iess_site_grouped$Family_grouped,
  levels = family_order
)
```

save & then read in the dataframes (for next time can just read in these!)
```{r}
write.csv(iess_sample.type_grouped, "iess_sample.type_grouped.csv")
write.csv(iess_site_grouped, "iess_site_grouped.csv")
```

Scaling colors for visual comparisons 
(all the colors on the same scale) 
```{r}
library(dplyr)
library(ggplot2)
library(scales)
library(RColorBrewer)

# Clean/standardize names
iess_sample.type_grouped$Family_grouped <- trimws(as.character(iess_sample.type_grouped$Family_grouped))
iess_site_grouped$Family_grouped <- trimws(as.character(iess_site_grouped$Family_grouped))


#all families 
all_families <- unique(c(
  iess_sample.type_grouped$Family_grouped,
  iess_site_grouped$Family_grouped
))

# color palette *want contrast*
my_colors <- colorRampPalette(
  brewer.pal(12, "Paired")
)(length(all_families))

my_colors <- setNames(my_colors, all_families)

# color specific categories
my_colors[names(my_colors) %in% "Assigned at Higher Taxonomy"] <- "darkblue"
my_colors[names(my_colors) %in% "Rare Families"]              <- "grey80"
my_colors[names(my_colors) %in% "Unknown Family"]             <- "grey0"


# Set factor levels (for consistency)
iess_sample.type_grouped$Family_grouped <- factor(
  iess_sample.type_grouped$Family_grouped,
  levels = names(my_colors)
)

iess_site_grouped$Family_grouped <- factor(
  iess_site_grouped$Family_grouped,
  levels = names(my_colors)

)
```

Figure 4B ~ site
```{r}
#cleaning and ordering Monthc column
iess_site_grouped$Site <- trimws(iess_site_grouped$Site)
unique(iess_site_grouped$Site)
iess_site_grouped$Site <- factor(
  iess_site_grouped$Site,
  levels = c( "TM", "SB")
)

#plotting IESS Site

plot2 <- ggplot(iess_site_grouped,
       aes(x = Site, y = Percentage, fill = Family_grouped)) +
  geom_bar(stat = "identity") +
  labs(x = "Site", y = "Relative Abundance (%)", fill = "Family") +
  theme_bw(base_size = 16) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    panel.border = element_blank()
  ) +
  scale_fill_manual(values = my_colors)

plot2


```

Figure 4B ~ sample.type
```{r}
#cleaning and ordering Monthc column
iess_sample.type_grouped$Sample.Type <- trimws(iess_sample.type_grouped$Sample.Type)
unique(iess_sample.type_grouped$Sample.Type)
iess_sample.type_grouped$Sample.Type<- factor(
 iess_sample.type_grouped$Sample.Type,
  levels = c( "Internal", "External")
)

#plotting IESS Site
plot3 <- ggplot(iess_sample.type_grouped,
       aes(x = Sample.Type, y = Percentage, fill = Family_grouped)) +
  geom_bar(stat = "identity") +
  labs(x = "Sample Type", y = "Relative Abundance (%)", fill = "Family") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    panel.border = element_blank()
  ) +
  scale_fill_manual(values = my_colors)

plot3
```
put side by side? 
```{r}
#remove legend frim plot 3 

plot3.noleg <- ggplot(iess_sample.type_grouped,
       aes(x = Sample.Type, y = Percentage, fill = Family_grouped)) +
  geom_bar(stat = "identity") +
  labs(x = "Sample Type", y = "Relative Abundance (%)", fill = "Family") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.margin = margin(5, 5, 5, 5),
    panel.grid = element_blank(),
    panel.border = element_blank()
  ) +
  scale_fill_manual(values = my_colors)

plot3.noleg
```

```{r}
library(cowplot)
# Extract legend
plot2.1 <- get_legend(plot2)

# Remove legend from original plot if desired
plot2.noleg <- plot2 +
  theme(legend.position = "none")

```

```{r}
#fix up side by side bar plot 
mid <- cowplot::plot_grid(
  plot2.noleg,
  plot3.noleg,
  plot2.1,
  ncol = 3,
  rel_widths = c(.8, .4, .4)
)
mid
```

figure 4a ~ ordination plot
```{r}
#finalpcoaplot_internalexternal
ie.pcoa.1 <- plot_ordination(iess.ps.cleaner.clr,ordinate(iess.ps.cleaner.clr,"PCoA", "euclidean"),color="Sample.Type", shape="Site")+
  stat_ellipse(linewidth=.5)+
  geom_point(size = 3, alpha = 0.8)+
  theme_classic(base_size = 18)+
  #ggtitle(substitute(paste(italic("Location"))))+
  scale_color_manual(name = "Sample.Type", values = c("palegreen4","tomato4"))
ie.pcoa.1

```

 

Final combo plot!

```{r}
iess.fin <- cowplot::plot_grid(ie.pcoa.1,mid, ncol = 2, rel_widths    = c(.5,1))
iess.fin
```
```{r}
ggsave("iess.fin.pdf",
       plot = iess.fin,
       width = 29,
       height = 12,
       dpi = 300)
```

*no pam for iess (because dependent internal/external samples with no mat id to pair them)
*no graphs for alpha or nutrients 

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~PS Stats!~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
start back here 







read in *need to upload these to github*
```{r}
sm.pam <- read.csv("SM.PAM copy.csv")
di.pam <- read.csv("DI.PAM copy.csv")
```


~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~DS Stats!~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~




