---
title: "SpacTempGithub"
output: html_document
date: "2026-05-13"
---

loading libraries
```{r}
library(stringr)
library(dplyr)
library(stats)
library(ggplot2)
library(ggpubr)
library(vegan)
library(cowplot)
library(tidyverse)
library(phyloseq)
library(microbiome)
library("funfuns")
library(pairwiseAdonis)
library(viridis)
library("microViz")
```

```{r}
getwd()
```

pipelines are as follows:
- for phenological (seasonal sampling) + diel sampling bioinformatic pipeline - file is titled "sm.di.pipeline.Rmd
- for IESS bioinformatic pipeline - file is titled "ie.pipeline.Rmd" 

uploading ps.cleaner (raw files) 
iess.ps.cleaner = ps.cleaner for IESS samples (internal/external spatial survey)
ps.ps.cleaner = ps.cleaner for PS samples (phenological)
di.ps.cleaner = ps.cleaner for DI samples (diel)

raw counts..need to clr tranform them
```{r}
iess.ps.cleaner <- readRDS("CH1_IE_16S_ps.cleaner.RDS")
ps.ps.cleaner <- readRDS("sm1.ps.cleaner.rds")
di.ps.cleaner <- readRDS("diel.ps.cleaner.rds")
```

ensuring raw counts
```{r}
otu_table(iess.ps.cleaner)[1:5, 1:5]
otu_table(ps.ps.cleaner)[1:5, 1:5]
otu_table(di.ps.cleaner)[1:5, 1:5]
```

clr transformation for each 
```{r}
iess.ps.cleaner.clr <- microbiome::transform(iess.ps.cleaner,'clr')
iess.seqtab.clr <- data.frame(otu_table(iess.ps.cleaner.clr))

ps.ps.cleaner.clr <- microbiome::transform(ps.ps.cleaner,'clr')
ps.seqtab.clr <- data.frame(otu_table(ps.ps.cleaner.clr))

di.ps.cleaner.clr <- microbiome::transform(di.ps.cleaner,'clr')
di.seqtab.clr <- data.frame(otu_table(di.ps.cleaner.clr))
```

ensuring clr tranformation and saving dataframes
```{r}
otu_table(iess.ps.cleaner.clr)[1:5, 1:5]
otu_table(ps.ps.cleaner.clr)[1:5, 1:5]
otu_table(di.ps.cleaner.clr)[1:5, 1:5]
```
good^
```{r}
saveRDS(iess.ps.cleaner.clr, "iess.ps.cleaner.clr.rds")
saveRDS(ps.ps.cleaner.clr, "ps.ps.cleaner.clr.rds")
saveRDS(di.ps.cleaner.clr, "di.ps.cleaner.clr.rds")
```
all saved & uploaded into github to start back here! 

Now let's start with the stats
~~~~~~~~~~~~~~~SURVEY STATS & Figures~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
Note: Figure 2 Map (not made in r, made on arcmap gis)

PCA 1 - GROUP BY MAT
```{r}
pca.df <- read.csv("pca.df.update.csv")
```

```{r}
# create factor for grouping
pca.df$mat_group <- factor(pca.df$mat, labels = c("No mats", "Mats"))
```

```{r}
# select only numeric environmental variables
env_vars <- pca.df %>%
  select(cond, sal, TDS, pH, RDO.Sat, temp, DOC)
```

```{r}
# run PCA (scale is important)
pca <- prcomp(env_vars, scale. = TRUE)
```

```{r}
biplot(pca)
```
Figure 3A - labels manually added later 
```{r}
library(factoextra)
pca_mat1.2 <- fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage = pca.df$mat_group,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  label = "none",   # <-- removes labels, keeps arrows
  col.var = "black",
  pointsize = 3,
  alpha.ind = 0.8,
  repel = TRUE,
  legend.title = "Mat Presence",
  title = "PCA of Water Quality Parameters",
  select.var = list(name = c("pH", "cond", "sal", "DOC", "temp", "RDO.Sat"))
) +
  theme_minimal(base_size = 14)

pca_mat1.2

```
Notes on PCA1: 

- Dim1 = 57.7% of the variation
- Dim2 = 22.2% of the variation
- together, almost 80% 

- Mats (teal triangles) cluster more toward the right side of Dim1
- No mats (red circles) are more spread out and include more of the left side
- mat presence is associated with a particular water chemistry signature, not random conditions

- Variables pointing to the RIGHT (positive Dim1): cond (conductivity), TDS, sal (salinity), pH (slightly up-right)

- These are tightly grouped  ~  positively correlated with each other
- So Dim1 is salinity / ionic strength gradient.

-Sites with mats tend to occur where: higher conductivity, higher TDS, higher salinity, slightly higher pH

- Variables pointing DOWN (negative Dim2): temp, RDO.Sat (dissolved oxygen saturation)
- So Dim2 is mostly a temperature / oxygen gradient.
- Lower Dim2 values = warmer water, lower DO saturation

- DOC arrow points left
- DOC is opposite of salinity/TDS/cond
- So:High DOC = low salinity & conductivity, Low DOC = higher salinity waters
- makes sense!

```{r}
#save as PDF
ggsave(
  filename = "pca.mat.nolabels.pdf",
  plot = pca_mat1.2,
  width = 8,
  height = 6
)
```

statistical test to back this up: PERMANOVA
```{r}
env_vars <- pca.df[, c("cond", "TDS", "sal", "pH", "DOC", "temp", "RDO.Sat")]
library(vegan)
env_scaled <- scale(env_vars)
env_dist <- dist(env_scaled, method = "euclidean")
adonis2(env_dist ~ mat_group, 
        data = pca.df,
        permutations = 9999)
```
not stat supported by a PERMANOVA boo hoo
beta dispersion assumption passes ~
```{r}
#BETADISPERSIONCHECK
disp1 <- betadisper(env_dist, pca.df$mat_group)
anova(disp1)
```
PCA 2 - GROUP BY MONTH!

```{r}
# create factor for grouping
pca.df$month_group <- factor(
  pca.df$month,
  levels = c("Mar_2023", "June_2023", "July_2023"),
  labels = c("March", "June", "July")
)

# select only numeric environmental variables
env_vars <- pca.df %>%
  select(cond, sal, TDS, pH, RDO.Sat, temp, DOC)

# run PCA (scale is important)
pca <- prcomp(env_vars, scale. = TRUE)
```

```{r}
biplot(pca)
```
Figure 3B ~ labels manually fixed later
```{r}
pca_month <- fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage = pca.df$month_group,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  label = "var",   
  col.var = "black",
  pointsize = 3,
  alpha.ind = 0.8,
  repel = TRUE,
  legend.title = "Month",
  title = "PCA of Water Quality Parameters",
  select.var = list(name = c("pH", "cond", "TDS", "sal", "DOC", "temp", "RDO.Sat"))
) +
  theme_minimal(base_size = 14)

pca_month

```
```{r}
# Optionally, save as PDF
ggsave(
  filename = "pca.month.nolabel.pdf",
  plot = pca_month,
  width = 8,
  height = 6
)
```


statistical test to back this up: PERMANOVA
```{r}
env_vars <- pca.df[, c("cond", "TDS", "sal", "pH", "DOC", "temp", "RDO.Sat")]
library(vegan)
env_scaled <- scale(env_vars)
env_dist <- dist(env_scaled, method = "euclidean")
adonis2(env_dist ~ month_group, 
        data = pca.df,
        permutations = 9999)

#BETADISPERSIONCHECK
disp2 <- betadisper(env_dist, pca.df$month_group)
anova(disp2)
#passes beta dispersion result!

#pariwise test
# install if needed
# install.packages("devtools")
# devtools::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")

library(pairwiseAdonis)

# Run pairwise PERMANOVA
pairwise.adonis2(env_dist ~ month_group, data = pca.df, permutations = 9999, p.adjust.m = "BH")

```
differences seen between 3 sampling months!

PCA 1 - GROUP BY WATER TYPE
```{r}
# create factor for grouping
pca.df$water_group <- factor(
  pca.df$water,
  levels = c("sound", "lagoon", "channel", "river estuary"),
  labels = c("S", "L", "C", "RE")
)
```

Figure 3C ~ labels manually added later
```{r}
pca_water <- fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage = pca.df$water_group,
  addEllipses = FALSE,
  #ellipse.level = 0.95,
  label = "none",   
  col.var = "black",
  pointsize = 3,
  alpha.ind = 0.8,
  repel = TRUE,
  legend.title = "Water Type",
  title = "",
  select.var = list(name = c("pH", "cond", "TDS", "sal", "DOC", "temp", "RDO.Sat"))
) +
  theme_minimal(base_size = 14)

pca_water
```

To save
```{r}
# Optionally, save as PDF
ggsave(
  filename = "pca.water.nolab.noellipses.pdf",
  plot = pca_water,
  width = 8,
  height = 6
)
```

```{r}
adonis2(env_dist ~ water_group, 
        data = pca.df,
        permutations = 9999)

#BETADISPERSIONCHECK
disp5 <- betadisper(env_dist, pca.df$water_group)
anova(disp5)

# Run pairwise PERMANOVA
#pairwise.adonis2(env_dist ~ month_group, data = pca.df, permutations = 9999, p.adjust.m = "BH")
```
*not stat sig different based on water type~ beta disp passes

GROUP BY REGION (NEARBY/WATER CLUMPING SPOTS)
```{r}
library(dplyr)

# create factor for geographic grouping
pca.df <- pca.df %>%
  mutate(site_group = case_when(
    site %in% c("HH") ~ "Western_Sound",
    site %in% c("JI") ~ "Eastern_Sound",
    site %in% c("MM", "SBE", "SBC", "SB") ~ "Central_Lagoon_Sound",
    site %in% c("TM", "BS", "FM") ~ "Northern_Lagoon_Sound_Channel",
    site %in% c("PI", "TI") ~ "Northern_River_Estuary",
    TRUE ~ "Other"
  ))

pca.df$site_group <- factor(pca.df$site_group,
                            levels = c("Western_Sound",
                                       "Eastern_Sound",
                                       "Central_Lagoon_Sound",
                                       "Northern_Lagoon_Sound_Channel",
                                       "Northern_River_Estuary"),
                            labels = c("WS",
                                       "ES",
                                       "CLS",
                                       "NLSC",
                                       "NRE"))


```
 
Figure 3D ~ labels manually added later
```{r}
pca_loc <- fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage = pca.df$site_group,
  addEllipses = FALSE,
  label = "none",   
  col.var = "black",
  pointsize = 3,
  alpha.ind = 0.8,
  repel = TRUE,
  legend.title = "region",
  title = "",
  select.var = list(name = c("pH", "cond", "TDS", "sal", "DOC", "temp", "RDO.Sat"))
) +
  theme_minimal(base_size = 14)

pca_loc
```
To save
```{r}
# Optionally, save as PDF
ggsave(
  filename = "pca.loc.nolab.noellipses.pdf",
  plot = pca_loc,
  width = 8,
  height = 6
)
```
 
stats check 
```{r}
adonis2(env_dist ~ site_group, 
        data = pca.df,
        permutations = 9999)

#BETADISPERSIONCHECK
disp6 <- betadisper(env_dist, pca.df$site_group)
anova(disp6)

# Run pairwise PERMANOVA
#pairwise.adonis2(env_dist ~ month_group, data = pca.df, permutations = 9999, p.adjust.m = "BH")
```
*not stat sig different based on region ~ beta disp passes

putting plot together ~ figure 3 final!



```{r}
library(cowplot)
pca.top <-cowplot::plot_grid(
   pca_mat1.2,
   pca_month,
  nrow = 1,
  align = "h",
  axis = "tb")
pca.top
```

```{r}
pca.bottom <- cowplot::plot_grid(
   pca_water,
   pca_loc,
  nrow = 1,
  align = "h",
  axis = "tb")
pca.bottom
```
```{r}
library(cowplot)
final.pca <- cowplot::plot_grid(
pca.top,
pca.bottom,
  nrow = 2,
  align = "h",
  axis = "tb")

```

```{r}
final.pca
```
```{r}
ggsave(
  filename = "final.pca.pdf",
  plot = final.pca,
  width = 10,
  height = 10,
  device = cairo_pdf
)

```


~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~IESS Stats!~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```{r}
iess.seqtab <- data.frame(otu_table(iess.ps.cleaner))
iess.dist.all <- vegdist(iess.seqtab, method = "euclidean")
```

extract metadata
```{r}
iess.samdf <- data.frame(sample_data(iess.ps.cleaner))
```

Beta Dispersion!
```{r}
iess.bet.all1 <- betadisper(iess.dist.all,iess.samdf$Sample.Type)
anova(iess.bet.all1) 

iess.bet.all2 <- betadisper(iess.dist.all,iess.samdf$Site)
anova(iess.bet.all2) 
```
Not significant p=.1002, .9643

Beta Diversity
```{r}
iess.samdf$Sample.Type <- as.factor(iess.samdf$Sample.Type)
iess.samdf$Site <- as.factor(iess.samdf$Site)
adonis2(iess.dist.all ~ Sample.Type * Site , data=iess.samdf, permutations=9999)
```
Alpha Diversity!
alpha diversity dataframe set up: 
how did we make these? 
alpha_diversity<- data.frame(estimate_richness(ps.cleaner, split=TRUE, measures=c("Shannon","InvSimpson","Observed")))
alpha_diversity$Evenness <- alpha_diversity$Shannon / log(alpha_diversity$Observed)

^did this for each experiment and saved dataframes to be uploaded below 
ps.cleaner phyloseq object should not have transformed data (should use raw or minimally processed data for this)
*if included below then the model passes assumptions + significant results 

```{r}
iess.alpha <- data.frame(estimate_richness(iess.ps.cleaner, split=TRUE, measures=c("Shannon","InvSimpson","Observed")))
iess.alpha$Evenness <- iess.alpha$Shannon / log(iess.alpha$Observed)

sample_metadata <- data.frame(sample_data(iess.ps.cleaner))
iess.alpha <- cbind(sample_metadata, iess.alpha)

write.csv(iess.alpha,"iess.alpha.csv")
```

*next time just read in iess.alpha
model for shannon, richness, evenness, invsimpson

which model type? 

the 1|Mat accounts for samples within mats not being truly independent, especially important for iess experiment
*only included stat sig models, but ran these for each alpha div stat
*no need for interactions between variables 
```{r}
iess.alpha.2 <- lmer(InvSimpson ~ Site + Sample.Type + (1|Mat) , data = iess.alpha, REML = FALSE)
summary(iess.alpha.2)
```
```{r}
# Plot residuals against fitted values
plot(residuals(iess.alpha.2) ~ fitted(iess.alpha.2))

# Shapiro-Wilk normality test
shapiro.test(residuals(iess.alpha.2))

# Q-Q plot of residuals
qqnorm(residuals(iess.alpha.2))
qqline(residuals(iess.alpha.2))
plot_model(iess.alpha.2, type = 'diag')
#passes assumptions

```
passes model assumptions

```{r}
iess.alpha.3 <- lmer(Evenness ~ Site + Sample.Type + (1|Mat) , data = iess.alpha, REML = FALSE)
summary(iess.alpha.3)
```

```{r}
# Plot residuals against fitted values
plot(residuals(iess.alpha.3) ~ fitted(iess.alpha.3))

# Shapiro-Wilk normality test
shapiro.test(residuals(iess.alpha.3))

# Q-Q plot of residuals
qqnorm(residuals(iess.alpha.3))
qqline(residuals(iess.alpha.3))
plot_model(iess.alpha.3, type = 'diag')
#passes assumptions

```
passes model assumptions

just to ensure which one is higher 
```{r}
library(ggplot2)

ggplot(iess.alpha, aes(x = Sample.Type, y = Evenness, color = Site)) +
  geom_boxplot(position = position_dodge(width = 0.8), alpha = 0.3) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8)) +
  theme_classic() +
  labs(x = "Sample Type", y = "Evenness")


ggplot(iess.alpha, aes(x = Sample.Type, y = InvSimpson, color = Site)) +
  geom_boxplot(position = position_dodge(width = 0.8), alpha = 0.3) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8)) +
  theme_classic() +
  labs(x = "Sample Type", y = "InvSimpson")
```
*internal evenness and inv simpson  is higher
*plots above not included in paper, purely for result visualization

Diff Abund! 
(1) sample type
```{r}
ie.diff.abund1 <- ancombc2(
  data = iess.ps.cleaner,      # your phyloseq object
  tax_level = "Family",      # or "Genus", etc.
  fix_formula = "Sample.Type",  # variable to test
  p_adj_method = "holm",
  group = "Sample.Type",        # group variable for pairwise tests
  struc_zero = TRUE,         # detects true absences
  neg_lb = TRUE,
  em_control = list(
    tol = 1e-5,
    max_iter = 100,
    sensitivity = TRUE
  ),
  alpha = 0.05,
  global = TRUE
)
```
```{r}
#structural zero results (need to clean up in excel)
View(ie.diff.abund1$zero_ind)
#write.csv(ie.diff.abund1$zero_ind, "ie.diff.abund1_strzeros.csv", row.names = TRUE)
#sensitivity results (scores are low, this is why we negate global and pairwise results and focus on the robust structural zero results above)
View(ie.diff.abund1$pseudo_sens_tab)
#write.csv(ie.diff.abund1$pseudo_sens_tab, "ie.diff.abund1_sens.csv", row.names = TRUE)
```

(1) site
```{r}
ie.diff.abund2 <- ancombc2(
  data = iess.ps.cleaner,      # your phyloseq object
  tax_level = "Family",      # or "Genus", etc.
  fix_formula = "Site",  # variable to test
  p_adj_method = "holm",
  group = "Site",        # group variable for pairwise tests
  struc_zero = TRUE,         # detects true absences
  neg_lb = TRUE,
  em_control = list(
    tol = 1e-5,
    max_iter = 100,
    sensitivity = TRUE
  ),
  alpha = 0.05,
  global = TRUE
)
```
```{r}
#structural zero results (need to clean up in excel)
View(ie.diff.abund2$zero_ind)
#write.csv(ie.diff.abund2$zero_ind, "ie.diff.abund2_strzeros.csv", row.names = TRUE)
#sensitivity results (scores are low, this is why we negate global and pairwise results and focus on the robust structural zero results above)
View(ie.diff.abund2$pseudo_sens_tab)
#write.csv(ie.diff.abund2$pseudo_sens_tab, "ie.diff.abund2_sens.csv", row.names = TRUE)
```


Nutrient Stats/correlations! iess.alpha (has nutrient data) 


~look at differences of TOM, TOC, TN between sample types (internal vs external samples) &  across sites

(1) TOM

# Organic Matter
mod_OM <- lmer(X..Organic.Matter ~ Sample.Type + Site + (1|Mat),
               data = iess.alpha)
#assumptions
# Plot residuals against fitted values
plot(residuals(mod_OM) ~ fitted(mod_OM))

# Shapiro-Wilk normality test
shapiro.test(residuals(mod_OM))

# Q-Q plot of residuals
qqnorm(residuals(mod_OM))
qqline(residuals(mod_OM))
plot_model(mod_OM, type = 'diag')
#does not pass so wilcoxin paired test
`

~ Using wilcoxin test for TOM because PAIRED internal/external samples & lme model did not pass assumptions 
```{r}
wilcox.test(X..Organic.Matter ~ Sample.Type,
            paired = TRUE,
            data = iess.alpha)

wilcox.test(X..Organic.Matter ~ Site,
            paired = TRUE,
            data = iess.alpha)
```
```{r}
#not adding to paper
ggplot(iess.alpha, aes(x = Site, y = X..Organic.Matter , color = Site)) +
  geom_boxplot(position = position_dodge(width = 0.8), alpha = 0.3) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8)) +
  theme_classic() +
  labs(x = "site", y = "tom")
```



(2) TN
```{r}
#TN
mod_N <- lmer(X..Nitrogen ~ Sample.Type + Site + (1|Mat),
               data = iess.alpha)
#assumptions
# Plot residuals against fitted values
plot(residuals(mod_N) ~ fitted(mod_N))

# Shapiro-Wilk normality test
shapiro.test(residuals(mod_N))

# Q-Q plot of residuals
qqnorm(residuals(mod_N))
qqline(residuals(mod_N))
plot_model(mod_N, type = 'diag')

```


```{r}
library(lmerTest)

mod_N <- lmer(X..Nitrogen ~ Sample.Type + Site + (1|Mat),
              data = iess.alpha)

summary(mod_N)
```

(2) TOC
```{r}
#TOC
mod_TOC <- lmer(X..Organic.Carbon ~ Sample.Type + Site + (1|Mat),
               data = iess.alpha)
#assumptions
# Plot residuals against fitted values
plot(residuals(mod_TOC) ~ fitted(mod_TOC))

# Shapiro-Wilk normality test
shapiro.test(residuals(mod_TOC))

# Q-Q plot of residuals
qqnorm(residuals(mod_TOC))
qqline(residuals(mod_TOC))
plot_model(mod_TOC, type = 'diag')

```
```{r}
mod_TOC <- lmer(X..Organic.Carbon ~ Sample.Type + Site + (1|Mat),
               data = iess.alpha)
summary(mod_TOC)
```
```{r}
#not adding to paper
ggplot(iess.alpha, aes(x = Sample.Type, y = X..Organic.Carbon, color = Site)) +
  geom_boxplot(position = position_dodge(width = 0.8), alpha = 0.3) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8)) +
  theme_classic() +
  labs(x = "sample type", y = "toc")
```

~nutrient correlations 
to see families across the entire experiment:
```{r}
#note:iessmdffam ~ calculated and saved below ..need to  read back in here
iessmdffam %>%
  group_by(Family) %>%
  summarise(Total_Abundance = sum(Abundance)) %>%
  arrange(desc(Total_Abundance))
```

These are the >2% most relative abundant families (percentage) 
```{r}
top_families <- iessmdffam %>%
  group_by(Family) %>%
  summarise(Abundance = sum(Abundance)) %>%
  mutate(Percentage = (Abundance / sum(Abundance)) * 100) %>%
  filter(Percentage > 2) %>%
  arrange(desc(Percentage))

top_families
```

Calculating correlation coefficients based on Pearson
```{r}
# Load libraries
library(dplyr)
library(tidyr)
library(ggplot2)

#iess.alpha
#iessmdffam

# Merge dataframes 
merged.iess <- merge(iess.alpha, iessmdffam, by = "Sample")

#Define top families (this is taxa w > 2% relative abundance) ---
specific_family <- c("Rhodobacteraceae", "Desulfocapsaceae", "Flavobacteriaceae",
                     "Saprospiraceae","Woeseiaceae","Pirellulaceae",
                     "Halieaceae","A4b", "Cyclobacteriaceae", "Desulfosarcinaceae", "Puniceicoccaceae")

# Filter for top families & remove NAs 
merged_clean <- merged.iess %>%
  filter(Family %in% specific_family) %>%
  drop_na(Abundance, X..Organic.Matter.x, X..Organic.Carbon.x, X..Nitrogen.x)

#  Pivot nutrients to long format 
corr_df_long <- merged_clean %>%
  pivot_longer(
    cols = c(X..Organic.Matter.x, X..Organic.Carbon.x, X..Nitrogen.x),
    names_to = "Variable",
    values_to = "Nutrient"
  )

# Calculate correlations + p-values per family × nutrient
corr_family <- corr_df_long %>%
  group_by(Family, Variable) %>%
  summarize(
    Correlation_Coefficient = ifelse(
      length(unique(Nutrient)) > 1,
      cor.test(Abundance, Nutrient, method = "pearson")$estimate,
      NA
    ),
    
    P_value = ifelse(
      length(unique(Nutrient)) > 1,
      cor.test(Abundance, Nutrient, method = "pearson")$p.value,
      NA
    ),
    
    .groups = "drop"
  )
#write.csv(corr_family, "iess.corr_family.csv")

```
~sig stars added later~
corr_fam uploaded to github for easy readability of sig correlations 


HEATMAP #decided not to add to final paper!
```{r}
#make family in alphebetical order
corr_family$Family <- factor(
  corr_family$Family,
  levels = sort(unique(corr_family$Family), decreasing = TRUE)
)

iessheatmap_topfamily <- ggplot(corr_family, aes(x = Variable, y = Family, fill = Correlation_Coefficient)) +
  geom_tile(color = "white") +
  
  scale_fill_gradientn(
  colours = c(
    "#A6CEE3",  # light blue
    "#1F78B4",  # dark blue
    "#B2DF8A",  # light green
    "#33A02C"   # dark green
  ),
  name = "Pearson's Correlation"
) +
  scale_y_discrete(position = "right") +
  labs(x = "Nutrients", y = "Top Families") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    panel.border = element_blank(),
    legend.position = "right"
  )

iessheatmap_topfamily
```
*results arent super interesting (only 1 stat) so scratch above heatmap for final paper
*corr_fam uploaded to github for easy readability of sig correlations (just in case) 



Bar plots!
Need to sum groups (ie + site)
```{r}
library(microshades)
iess.ps.cleaner.rel <- transform_sample_counts(iess.ps.cleaner, function(x) x / sum(x))
fam.aggreg1 <- tax_glom(iess.ps.cleaner.rel, "Family")
iessmdffam<- prep_mdf(fam.aggreg1, subgroup_level = "Family")
```
```{r}
write.csv(iessmdffam, "iess.mdffam.csv")
```

site example below: 
*repeated for sample.type 
```{r}
library(dplyr)
iessfam_site <- iessmdffam %>%
  group_by(Site, Family) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

#convert to relative abundance within each transect
iessfam_site <- iessfam_site %>%
  group_by(Site) %>%
  mutate(Percentage = (Abundance / sum(Abundance)) * 100)

#grouping rules (same as above) applied
#families less than 1% relatively abundant grouped in "rare families" category 
#families only assigned at higher taxonomic rankings (phyla, class, order, ect) grouped into "Assigned at Higher Taxonomy" category
#note: some families were "unknown" meaning silva database was unable to find a match, so there should also be an "unknown" category unless all samples had identification to the family level

iessfam_site$Family_grouped <- ifelse(
  grepl("_", iessfam_site$Family),
  "Assigned at Higher Taxonomy",
  ifelse(iessfam_site$Percentage < 1, "Rare Families", iessfam_site$Family)
)

#re-sum after grouping (bc grouped categories are now collapsed)
iess_site_grouped <- iessfam_site %>%
  group_by(Site, Family_grouped) %>%
  summarise(Percentage = sum(Percentage), .groups = "drop")

#consistent stacking order based on overall abundance!
family_order <- iess_site_grouped %>%
  group_by(Family_grouped) %>%
  summarise(Total = sum(Percentage)) %>%
  arrange(Total) %>%
  pull(Family_grouped)

iess_site_grouped$Family_grouped <- factor(
  iess_site_grouped$Family_grouped,
  levels = family_order
)
```

save & then read in the dataframes (for next time can just read in these!)
```{r}
write.csv(iess_sample.type_grouped, "iess_sample.type_grouped.csv")
write.csv(iess_site_grouped, "iess_site_grouped.csv")
```

Scaling colors for visual comparisons 
(all the colors on the same scale) 
```{r}
library(dplyr)
library(ggplot2)
library(scales)
library(RColorBrewer)

# Clean/standardize names
iess_sample.type_grouped$Family_grouped <- trimws(as.character(iess_sample.type_grouped$Family_grouped))
iess_site_grouped$Family_grouped <- trimws(as.character(iess_site_grouped$Family_grouped))


#all families 
all_families <- unique(c(
  iess_sample.type_grouped$Family_grouped,
  iess_site_grouped$Family_grouped
))

# color palette *want contrast*
my_colors <- colorRampPalette(
  brewer.pal(12, "Paired")
)(length(all_families))

my_colors <- setNames(my_colors, all_families)

# color specific categories
my_colors[names(my_colors) %in% "Assigned at Higher Taxonomy"] <- "darkblue"
my_colors[names(my_colors) %in% "Rare Families"]              <- "grey80"
my_colors[names(my_colors) %in% "Unknown Family"]             <- "grey0"


# Set factor levels (for consistency)
iess_sample.type_grouped$Family_grouped <- factor(
  iess_sample.type_grouped$Family_grouped,
  levels = names(my_colors)
)

iess_site_grouped$Family_grouped <- factor(
  iess_site_grouped$Family_grouped,
  levels = names(my_colors)

)
```

Figure 4B ~ site
```{r}
#cleaning and ordering Monthc column
iess_site_grouped$Site <- trimws(iess_site_grouped$Site)
unique(iess_site_grouped$Site)
iess_site_grouped$Site <- factor(
  iess_site_grouped$Site,
  levels = c( "TM", "SB")
)

#plotting IESS Site

plot2 <- ggplot(iess_site_grouped,
       aes(x = Site, y = Percentage, fill = Family_grouped)) +
  geom_bar(stat = "identity") +
  labs(x = "Site", y = "Relative Abundance (%)", fill = "Family") +
  theme_bw(base_size = 16) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    panel.border = element_blank()
  ) +
  scale_fill_manual(values = my_colors)

plot2


```

Figure 4B ~ sample.type
```{r}
#cleaning and ordering Monthc column
iess_sample.type_grouped$Sample.Type <- trimws(iess_sample.type_grouped$Sample.Type)
unique(iess_sample.type_grouped$Sample.Type)
iess_sample.type_grouped$Sample.Type<- factor(
 iess_sample.type_grouped$Sample.Type,
  levels = c( "Internal", "External")
)

#plotting IESS Site
plot3 <- ggplot(iess_sample.type_grouped,
       aes(x = Sample.Type, y = Percentage, fill = Family_grouped)) +
  geom_bar(stat = "identity") +
  labs(x = "Sample Type", y = "Relative Abundance (%)", fill = "Family") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    panel.border = element_blank()
  ) +
  scale_fill_manual(values = my_colors)

plot3
```
put side by side? 
```{r}
#remove legend frim plot 3 

plot3.noleg <- ggplot(iess_sample.type_grouped,
       aes(x = Sample.Type, y = Percentage, fill = Family_grouped)) +
  geom_bar(stat = "identity") +
  labs(x = "Sample Type", y = "Relative Abundance (%)", fill = "Family") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.margin = margin(5, 5, 5, 5),
    panel.grid = element_blank(),
    panel.border = element_blank()
  ) +
  scale_fill_manual(values = my_colors)

plot3.noleg
```

```{r}
library(cowplot)
# Extract legend
plot2.1 <- get_legend(plot2)

# Remove legend from original plot if desired
plot2.noleg <- plot2 +
  theme(legend.position = "none")

```

```{r}
#fix up side by side bar plot 
mid <- cowplot::plot_grid(
  plot2.noleg,
  plot3.noleg,
  plot2.1,
  ncol = 3,
  rel_widths = c(.8, .4, .4)
)
mid
```

figure 4a ~ ordination plot
```{r}
#finalpcoaplot_internalexternal
ie.pcoa.1 <- plot_ordination(iess.ps.cleaner.clr,ordinate(iess.ps.cleaner.clr,"PCoA", "euclidean"),color="Sample.Type", shape="Site")+
  stat_ellipse(linewidth=.5)+
  geom_point(size = 3, alpha = 0.8)+
  theme_classic(base_size = 18)+
  #ggtitle(substitute(paste(italic("Location"))))+
  scale_color_manual(name = "Sample.Type", values = c("palegreen4","tomato4"))
ie.pcoa.1

```

 

Final combo plot!

```{r}
iess.fin <- cowplot::plot_grid(ie.pcoa.1,mid, ncol = 2, rel_widths    = c(.5,1))
iess.fin
```
```{r}
ggsave("iess.fin.pdf",
       plot = iess.fin,
       width = 29,
       height = 12,
       dpi = 300)
```

*no pam for iess (because dependent internal/external samples with no mat id to pair them)
*no graphs for alpha or nutrients 

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~PS Stats!~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

*fix month order - make it chronological in time!!!!!!!

```{r}
ps.seqtab <- data.frame(otu_table(ps.ps.cleaner))
ps.dist.all <- vegdist(ps.seqtab, method = "euclidean")
```

extract metadata
```{r}
ps.samdf <- data.frame(sample_data(ps.ps.cleaner))
```

update sample dataframe

```{r}
ps.samdf <- ps.samdf %>%
  mutate(
    Month_Year = paste(Month, Year, sep = "_"),
    Season = case_when(
      Month %in% c("March", "April", "May") ~ "Spring",
      Month %in% c("June", "July", "August") ~ "Summer",
      Month %in% c("September", "October", "November") ~ "Fall",
      Month %in% c("December", "January", "February") ~ "Winter",
      TRUE ~ NA_character_
    )
  )
```

Beta Dispersion!
```{r}
ps.bet.all1 <- betadisper(ps.dist.all,ps.samdf$Month_Year)
anova(ps.bet.all1) 
tukey1 <- TukeyHSD(ps.bet.all1)
tukey1

ps.bet.all2 <- betadisper(ps.dist.all,ps.samdf$Season)
anova(ps.bet.all2) 
tukey2 <- TukeyHSD(ps.bet.all2)
tukey2

ps.bet.all3 <- betadisper(ps.dist.all,ps.samdf$Site)
anova(ps.bet.all3) 

```
Not significant p=.1002, .9643

Beta Diversity
```{r}
ps.samdf$Month_Year <- as.factor(ps.samdf$Month_Year)
ps.samdf$Season <- as.factor(ps.samdf$Season)
ps.samdf$Site <- as.factor(ps.samdf$Site)
adonis2(ps.dist.all ~ Month_Year * Site , data=ps.samdf, permutations=9999) #this one is more informative
adonis2(ps.dist.all ~ Season * Site , data=ps.samdf, permutations=9999)
```
~ focusing on month_year * site interaction 
~ site-stratified permanova

```{r}
sites <- unique(ps.samdf$Site)

for (s in sites) {

  cat("\n--- PERMANOVA: Month_Year differences within Site =", s, "---\n")

  site_idx <- ps.samdf$Site == s

  dist_subset <- as.dist(as.matrix(ps.dist.all)[site_idx, site_idx])
  meta_subset <- ps.samdf[site_idx, ]

  if (length(unique(meta_subset$Month_Year)) > 1) {
    result <- adonis2(dist_subset ~ Month_Year,
                      data = meta_subset,
                      permutations = 9999)
    print(result)
  } else {
    cat("Not enough Month_Year levels for this site.\n")
  }
}
```

*yes, month_years differ at each site...but which ones? maybe flop for more specifics

does sites differ within each month_year? 
```{r}
months <- unique(ps.samdf$Month_Year)

for (m in months) {

  cat("\n--- PERMANOVA: Site differences within Month_Year =", m, "---\n")

  meta_subset <- subset(ps.samdf, Month_Year == m)

  # subset distance matrix using matching rownames
  dist_subset <- as.dist(
    as.matrix(ps.dist.all)[rownames(ps.samdf) %in% rownames(meta_subset),
                           rownames(ps.samdf) %in% rownames(meta_subset)]
  )

  meta_subset <- meta_subset[match(attr(dist_subset, "Labels"), rownames(meta_subset)), ]

  if (length(unique(meta_subset$Site)) > 1) {

    result <- adonis2(dist_subset ~ Site,
                      data = meta_subset,
                      permutations = 9999)

    print(result)

  } else {
    cat("Not enough Site levels for this Month_Year.\n")
  }
}
```
~~~~~~~~~~~~~~~~~~~~~Put plots in chronological order and cowplot them!!
Community Comp Plots: 
```{r}
sd <- sample_data(ps.ps.cleaner.clr)

sd$Month_Year <- factor(
  paste(sd$Month, sd$Year, sep = "_"),
  levels = c(
    "June_2023", "July_2023","October_2023", "December_2023", "February_2024", "April_2024","June_2024", "August_2024", "July_2025" )
)

sd$Site <- as.factor(sd$Site)

sample_data(ps.ps.cleaner.clr) <- sd
```

```{r}
unique(sample_data(ps.ps.cleaner.clr)$Month_Year)
```

```{r}
sm.pcaochsht.1 <- plot_ordination(
  ps.ps.cleaner.clr,
  ordinate(ps.ps.cleaner.clr, "PCoA", "euclidean"),
  shape = "Site",
  color = "Month_Year"
) +
  geom_point(size = 6, alpha = 0.8) +
  scale_color_viridis_d(name = "Month_Year") +
  theme_classic(base_size = 22)

sm.pcaochsht.1
```

```{r}
sm.pcaochsht.1 <- plot_ordination(
  ps.ps.cleaner.clr,
  ordinate(ps.ps.cleaner.clr, "PCoA", "euclidean"),
  shape = "Site",
  color = "Month_Year"
) +
  geom_point(size = 6, alpha = 0.8) +
  stat_ellipse(aes(group = Site), type = "t", linewidth = .2) +
  scale_color_viridis_d(name = "Month_Year") +
  theme_classic(base_size = 22) +
  theme(
    axis.title = element_text(size = 14),
    axis.text  = element_text(size = 12)
  )


sm.pcaochsht.1 <- sm.pcaochsht.1 +
  theme(
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 10),

    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 10),
    legend.key.size = unit(0.4, "cm")
  )

sm.pcaochsht.1
```

Interaction Plots
Month_Year*Site interaction ~ to better visualize this
```{r}
ord.df1 <- read.csv("sm.ordination_df copy.csv")
ord.df2 <- read.csv("sm.ordination_df2 copy.csv")
```

```{r}
month_levels <- c(     "June_2023", "July_2023","October_2023", "December_2023", "February_2024", "April_2024","June_2024", "August_2024", "July_2025" )
ps.int.1 <- ggplot(
  ord.df1,
  aes(
    x = factor(Site),
    y = PCoA1,
    color = factor(Month_Year, levels = month_levels),
    group = Month_Year
  )
) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  scale_color_viridis_d(name = "Month_Year") +
  labs(
    title = "Interaction plot",
    x = "Site",
    y = "PCoA Axis 1 score"
  ) +
  theme_minimal()

ps.int.1
```

```{r}
ps.int.2 <- ggplot(
  ord.df1,
  aes(
    x = factor(Month_Year, levels = month_levels),
    y = PCoA1,
    color = Site,
    group = Site
  )
) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  scale_color_viridis_d(name = "Site") +
  labs(
    title = "Interaction plot",
    x = "Month_Year",
    y = "PCoA Axis 1 score"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.title.x = element_text(size = 10),
    axis.ticks.x = element_line(linewidth = 0.5),
    plot.margin = margin(10, 10, 10, 10)
  )

ps.int.2
```
*good ~ for final plot, top is pcoa, these 2 in the middle, bar plot (below) on bottom 


Bar plots! **START BACK UP HERE TOMORROW 

Need to sum groups (month_year)
```{r}
library(microshades)
ps.ps.cleaner.rel <- transform_sample_counts(ps.ps.cleaner, function(x) x / sum(x))
fam.aggreg1 <- tax_glom(ps.ps.cleaner.rel, "Family")
psmdffam<- prep_mdf(fam.aggreg1, subgroup_level = "Family")
```


```{r}

psmdffam <- psmdffam %>%
  mutate(
    Month_Year = paste(Month, Year, sep = "_"),
    Season = case_when(
      Month %in% c("March", "April", "May") ~ "Spring",
      Month %in% c("June", "July", "August") ~ "Summer",
      Month %in% c("September", "October", "November") ~ "Fall",
      Month %in% c("December", "January", "February") ~ "Winter",
      TRUE ~ NA_character_
    )
  )
write.csv(psmdffam, "ps.mdffam.csv") #upload to github
```



month_year example below: 
*repeated for season & site
```{r}
library(dplyr)
pssfam_moyr <- psmdffam %>%
  group_by(Month_Year, Family) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

#convert to relative abundance within each transect
pssfam_moyr <- pssfam_moyr %>%
  group_by(Month_Year) %>%
  mutate(Percentage = (Abundance / sum(Abundance)) * 100)

#grouping rules (same as above) applied
#families less than 1% relatively abundant grouped in "rare families" category 
#families only assigned at higher taxonomic rankings (phyla, class, order, ect) grouped into "Assigned at Higher Taxonomy" category
#note: some families were "unknown" meaning silva database was unable to find a match, so there should also be an "unknown" category unless all samples had identification to the family level

pssfam_moyr$Family_grouped <- ifelse(
  grepl("_", pssfam_moyr$Family),
  "Assigned at Higher Taxonomy",
  ifelse(pssfam_moyr$Percentage < 1, "Rare Families", pssfam_moyr$Family)
)

#re-sum after grouping (bc grouped categories are now collapsed)
ps_moyr_grouped <- pssfam_moyr %>%
  group_by(Month_Year, Family_grouped) %>%
  summarise(Percentage = sum(Percentage), .groups = "drop")

#consistent stacking order based on overall abundance!
family_order <-  ps_moyr_grouped%>%
  group_by(Family_grouped) %>%
  summarise(Total = sum(Percentage)) %>%
  arrange(Total) %>%
  pull(Family_grouped)

ps_moyr_grouped$Family_grouped <- factor(
  ps_moyr_grouped$Family_grouped,
  levels = family_order
)
```

```{r}
library(dplyr)
pssfam_seas <- psmdffam %>%
  group_by(Season, Family) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

#convert to relative abundance within each transect
pssfam_seas <- pssfam_seas %>%
  group_by(Season) %>%
  mutate(Percentage = (Abundance / sum(Abundance)) * 100)

#grouping rules (same as above) applied
#families less than 1% relatively abundant grouped in "rare families" category 
#families only assigned at higher taxonomic rankings (phyla, class, order, ect) grouped into "Assigned at Higher Taxonomy" category
#note: some families were "unknown" meaning silva database was unable to find a match, so there should also be an "unknown" category unless all samples had identification to the family level

pssfam_seas$Family_grouped <- ifelse(
  grepl("_", pssfam_seas$Family),
  "Assigned at Higher Taxonomy",
  ifelse(pssfam_seas$Percentage < 1, "Rare Families", pssfam_seas$Family)
)

#re-sum after grouping (bc grouped categories are now collapsed)
ps_seas_grouped <- pssfam_seas %>%
  group_by(Season, Family_grouped) %>%
  summarise(Percentage = sum(Percentage), .groups = "drop")

#consistent stacking order based on overall abundance!
family_order <- ps_seas_grouped %>%
  group_by(Family_grouped) %>%
  summarise(Total = sum(Percentage)) %>%
  arrange(Total) %>%
  pull(Family_grouped)

ps_seas_grouped$Family_grouped <- factor(
  ps_seas_grouped$Family_grouped,
  levels = family_order
)
```

```{r}
library(dplyr)
pssfam_site <- psmdffam %>%
  group_by(Site, Family) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

#convert to relative abundance within each transect
pssfam_site <- pssfam_site %>%
  group_by(Site) %>%
  mutate(Percentage = (Abundance / sum(Abundance)) * 100)

#grouping rules (same as above) applied
#families less than 1% relatively abundant grouped in "rare families" category 
#families only assigned at higher taxonomic rankings (phyla, class, order, ect) grouped into "Assigned at Higher Taxonomy" category
#note: some families were "unknown" meaning silva database was unable to find a match, so there should also be an "unknown" category unless all samples had identification to the family level

pssfam_site$Family_grouped <- ifelse(
  grepl("_", pssfam_site$Family),
  "Assigned at Higher Taxonomy",
  ifelse(pssfam_site$Percentage < 1, "Rare Families", pssfam_site$Family)
)

#re-sum after grouping (bc grouped categories are now collapsed)
ps_site_grouped <- pssfam_site %>%
  group_by(Site, Family_grouped) %>%
  summarise(Percentage = sum(Percentage), .groups = "drop")

#consistent stacking order based on overall abundance!
family_order <- ps_site_grouped %>%
  group_by(Family_grouped) %>%
  summarise(Total = sum(Percentage)) %>%
  arrange(Total) %>%
  pull(Family_grouped)

ps_site_grouped$Family_grouped <- factor(
  ps_site_grouped$Family_grouped,
  levels = family_order
)
```

save & then read in the dataframes (for next time can just read in these!)
```{r}
write.csv(ps_moyr_grouped, "ps_moyr_grouped.csv")
write.csv(ps_seas_grouped, "ps_seas_grouped.csv")
write.csv(ps_site_grouped, "ps_site_grouped.csv")
```


Scaling colors for visual comparisons 
(all the colors on the same scale) 
```{r}
library(dplyr)
library(ggplot2)
library(viridisLite)

# Clean/standardize names
ps_moyr_grouped$Family_grouped <- trimws(as.character(ps_moyr_grouped$Family_grouped))
ps_seas_grouped$Family_grouped <- trimws(as.character(ps_seas_grouped$Family_grouped))
ps_site_grouped$Family_grouped <- trimws(as.character(ps_site_grouped$Family_grouped))


# Get all unique families across datasets
all_families <- sort(unique(c(
  ps_moyr_grouped$Family_grouped,
  ps_seas_grouped$Family_grouped,
  ps_site_grouped$Family_grouped
)))

# Create viridis color palette
my_colors <- setNames(
  viridisLite::viridis(
    n = length(all_families),
    option = "D"
  ),
  all_families

)


# Override colors for special categories
my_colors["Assigned at Higher Taxonomy"] <- "darkblue"
my_colors["Rare Families"] <- "grey80"
my_colors["Unknown Family"] <- "black"

ps_moyr_grouped$Family_grouped <- factor(
  ps_moyr_grouped$Family_grouped,
  levels = all_families
)

ps_seas_grouped$Family_grouped <- factor(
  ps_seas_grouped$Family_grouped,
  levels = all_families
)

ps_site_grouped$Family_grouped <- factor(
  ps_site_grouped$Family_grouped,
  levels = all_families
)

```


```{r}
ps_moyr_grouped$Month_Year <- factor(
  ps_moyr_grouped$Month_Year,
  levels = month_levels
)

ps.plot1 <- ggplot(
  ps_moyr_grouped,
  aes(x = Month_Year, y = Percentage, fill = Family_grouped)
) +
  geom_bar(stat = "identity") +
  labs(
    x = "Sampling Period (Month & Year)",
    y = "Relative Abundance (%)",
    fill = "Family"
  ) +
  scale_fill_manual(values = my_colors) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    panel.border = element_blank()
  )


ps.plot1 <- ps.plot1 +
  theme(
    legend.position = "left",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    legend.key.size = unit(0.5, "cm"),
    legend.spacing.y = unit(0.05, "cm")
  )

ps.plot1 <- ps.plot1 +
  guides(fill = guide_legend(ncol = 2))

ps.plot1
```

```{r}
#not going to use this one
seas_levels <- c(
  "Summer",
  "Fall",
  "Winter",
  "Spring"
)

ps_seas_grouped$Season <- factor(
  ps_seas_grouped$Season,
  levels = seas_levels
)

ps.plot2 <- ggplot(
  ps_seas_grouped,
  aes(x = Season, y = Percentage, fill = Family_grouped)
) +
  geom_bar(stat = "identity") +
  labs(
    x = "Season",
    y = "Relative Abundance (%)",
    fill = "Family"
  ) +
  scale_fill_manual(values = my_colors) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(colour = "black")
  )
ps.plot2
```

```{r}
#removed legend
site_levels <- c(
  "SB",
  "TM"
)

ps_site_grouped$Site <- factor(
  ps_site_grouped$Site,
  levels = site_levels
)

ps.plot3 <- ggplot(
  ps_site_grouped,
  aes(x = Site, y = Percentage, fill = Family_grouped)
) +
  geom_bar(stat = "identity") +
  labs(
    x = "Site",
    y = "Relative Abundance (%)",
    fill = "Family"
  ) +
  scale_fill_manual(values = my_colors) +
  theme_bw() +
  theme(
  panel.grid = element_blank(),
  panel.border = element_blank(),
  axis.line = element_line(colour = "black"),
  legend.position = "none"
)
ps.plot3
```


Combo composition plot ~ will do later
sm.pcaochsht.1
ps.int.1
ps.int.2
ps.plot1
ps.plot3

```{r}
library(cowplot)

ps.mid <- cowplot::plot_grid(
  ps.int.1,
  ps.int.2,
  ncol = 2,
  align = "h"
  #labels = c("A", "B")   # optional panel labels
)
ps.mid
```

```{r}
ps.bottom <- cowplot::plot_grid(
  ps.plot1,
  ps.plot3,
  ncol = 2,
  align = "h",
  rel_widths = c(1, .45))  # optional panel labels

ps.bottom
```

```{r}
ps.comp.plot <- cowplot::plot_grid(
  sm.pcaochsht.1,
  ps.mid,
  ps.bottom,
  ncol = 1,
  align = "h",
  rel_heights  = c(.5,1,1))  # optional panel labels

ps.comp.plot
```
```{r}
ggsave("ps_comp_plot.pdf", plot = ps.comp.plot, width = 13, height = 21, dpi = 300)
```


Diff Abund! 
(1) Month_Year
*repeat for:
(2) Season *change group variable in code below and rerun
(3) Site *change group variable in code below and rerun

```{r}
ps.diff.abund1 <- ancombc2(
  data = ps.ps.cleaner,      # your phyloseq object
  tax_level = "Family",      # or "Genus", etc.
  fix_formula = "Month_Year",  # variable to test
  p_adj_method = "holm",
  group = "Month_Year",        # group variable for pairwise tests
  struc_zero = TRUE,         # detects true absences
  neg_lb = TRUE,
  em_control = list(
    tol = 1e-5,
    max_iter = 100,
    sensitivity = TRUE
  ),
  alpha = 0.05,
  global = TRUE
)
```
```{r}
#structural zero results (need to clean up in excel)
View(ps.diff.abund1$zero_ind)
#write.csv(ps.diff.abund1$zero_ind, "ps.diff.abund1_strzeros.csv", row.names = TRUE)
#sensitivity results (scores are low, this is why we negate global and pairwise results and focus on the robust structural zero results above)
View(ps.diff.abund1$pseudo_sens_tab)
#write.csv(ps.diff.abund1$pseudo_sens_tab, "ps.diff.abund1_sens.csv", row.names = TRUE)
```

Alpha Diversity

alpha diversity dataframe set up: 
how did we make these? 
alpha_diversity<- data.frame(estimate_richness(ps.cleaner, split=TRUE, measures=c("Shannon","InvSimpson","Observed")))
alpha_diversity$Evenness <- alpha_diversity$Shannon / log(alpha_diversity$Observed)

^did this for each experiment and saved dataframes to be uploaded below 
ps.cleaner phyloseq object should not have transformed data (should use raw or minimally processed data for this)

*if included below then the model passes assumptions + significant results 

```{r}
ps.ps.alpha<- read.csv("alpha.div.sm copy.csv") #this one is uploaded to github 
```

setting factors
```{r}
ps.ps.alpha <- ps.ps.alpha %>%
  mutate(
    Month_Year = factor(Month_Year, levels = month_levels),
    Site = factor(Site)
  )
```

Shannon
```{r}
library(vegan)
ps.alpha.1 <-aov(Shannon ~ Site * Month_Year, data = ps.ps.alpha)
summary(ps.alpha.1)
```

```{r}
#Assumption checks
#1. Normality
qqnorm(residuals(ps.alpha.1))
qqline(residuals(ps.alpha.1), col = "red")
shapiro_test <- shapiro.test(residuals(ps.alpha.1))
print(shapiro_test)
#2. Homogeneity
library(car)
levene_test <- leveneTest(Shannon ~ Site * Month_Year, data = ps.ps.alpha)
print(levene_test)
#3 linearity
plot(ps.alpha.1, which = 1)
plot
#4 independence (think about sample design)
durbinWatsonTest(ps.alpha.1)
```

*did not pass assumptions, so do KW

we have to do independent nonparametric kruskal wallis tests...... cannot look at the interaction between site and month! 
```{r}
kruskal.test(Shannon ~ Month_Year, data = ps.ps.alpha)  # main effect of month
kruskal.test(Shannon ~ Site, data = ps.ps.alpha)        # main effect of site
```
Yes to month_year!

checking for spreads or extreme skews...
```{r}
boxplot(Shannon ~ Month_Year, data = ps.ps.alpha)
boxplot(Shannon ~ Site, data = ps.ps.alpha)
#Shannon diversity distributions differ among months (rank-based differences)
```
looks fine to me! ~ let's do post hoc test for month_year
```{r}
library(FSA)
dunn_results_ps1 <- dunnTest(Shannon ~ Month_Year,
                         data = ps.ps.alpha,
                         method = "bh")   # adjust p-values
dunn_results_ps1
```
specific months differ ~ what does this look like? 
April_2024 - August_2024	-2.96507744	0.0030260675	0.03631281	
August_2024 - February_2024	3.43702748	0.0005881359	0.01058645	
August_2024 - October_2023	3.48832640	0.0004860543	0.01749796	
(aug diversity is higher)


```{r}
# month plot for shannon diversity
ps.ps.alpha$Month_Year <- factor(
  ps.ps.alpha$Month_Year,
  levels = month_levels
)

library(ggpubr)
library(ggplot2)
library(viridisLite)

# more vibrant viridis palette for months
month_colors <- setNames(
  viridis(
    n = length(month_levels),
    option = "turbo"   # <- much more vivid than "D"
  ),
  month_levels
)

ps.shannon <- ggplot(
  ps.ps.alpha,
  aes(
    x = factor(Month_Year, levels = month_levels),
    y = Shannon,
    fill = factor(Month_Year, levels = month_levels)
  )
) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "black") +
  scale_fill_manual(values = month_colors) +
  coord_cartesian(ylim = c(4.5, 7.5)) +
  ylab("Shannon Diversity") +
  xlab("Sampling Period") +
  theme_classic(base_size = 14) +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 15),
    legend.position = "none"
  ) +
  stat_compare_means(
    comparisons = list(c("1", "5"), c("3", "5")),
    method = "t.test",
    label = "p.signif",
    label.y = c(7.3, 7.5)
  )

ps.shannon
```

Inv Simpson
```{r}
library(vegan)
ps.alpha.2 <-aov(InvSimpson ~ Site * Month_Year, data = ps.ps.alpha)
summary(ps.alpha.2)

#Assumption checks
#1. Normality
qqnorm(residuals(ps.alpha.2))
qqline(residuals(ps.alpha.2), col = "red")
shapiro_test <- shapiro.test(residuals(ps.alpha.2))
print(shapiro_test)
#2. Homogeneity
library(car)
levene_test <- leveneTest(InvSimpson ~ Site * Month_Year, data = ps.ps.alpha)
print(levene_test)
#3 linearity
plot(ps.alpha.2, which = 1)
plot
#4 independence (think about sample design)
durbinWatsonTest(ps.alpha.2)
#The Durbin-Watson statistic ranges from 0 to 4.
#2 = no autocorrelation
#< 2 = positive autocorrelation
#> 2 = negative autocorrelation
```
passes assumptions:

post hoc (month_year)
```{r}
TukeyHSD(ps.alpha.2, which = "Site")
TukeyHSD(ps.alpha.2, which = "Month_Year")
```
Inv Simpson: both site and month_year differ ~ but month_year does not have ind. months that differ
look at effect size:
```{r}
library(effectsize)
eta_squared(ps.alpha.2)
```
*both effect sizes are non-trivial
*will report site and month_year stat sig results for inv simpson


Richness
```{r}
library(vegan)
ps.alpha.3 <-aov(Observed ~ Site * Month_Year, data = ps.ps.alpha)
summary(ps.alpha.3)

#Assumption checks
#1. Normality
qqnorm(residuals(ps.alpha.3))
qqline(residuals(ps.alpha.3), col = "red")
shapiro_test <- shapiro.test(residuals(ps.alpha.3))
print(shapiro_test)
#2. Homogeneity
library(car)
levene_test <- leveneTest(Observed ~ Site * Month_Year, data = ps.ps.alpha)
print(levene_test)
#3 linearity
plot(ps.alpha.3, which = 1)
plot
#4 independence (think about sample design)
durbinWatsonTest(ps.alpha.3)
#The Durbin-Watson statistic ranges from 0 to 4.
#2 = no autocorrelation
#< 2 = positive autocorrelation
#> 2 = negative autocorrelation
```
```{r}
TukeyHSD(ps.alpha.3, which = "Month_Year")
```
august is driving differences!!!!
also differences between months in july 
August_2024-June_2023        1150.238681   470.04850 1830.428866 0.0000349
August_2024-June_2024        1179.700000   624.32704 1735.072961 0.0000002
August_2024-July_2023        1454.238681   774.04850 2134.428866 0.0000001
August_2024-July_2025         727.538681    86.24922 1368.828137 0.0150343
October_2023-July_2025       -715.461319 -1356.75078  -74.171863 0.0179395
February_2024-July_2025      -735.661319 -1376.95078  -94.371863 0.0133311
April_2024-July_2025         -645.061319 -1286.35078   -3.771863 0.0475876
October_2023-August_2024    -1443.000000 -1998.37296 -887.627039 0.0000000 (p < 0.01)
December_2023-August_2024   -1301.915409 -1872.50688 -731.323933 0.0000000 (p < 0.01)
February_2024-August_2024   -1463.200000 -2018.57296 -907.827039 0.0000000 (p < 0.01)
April_2024-August_2024      -1372.600000 -1927.97296 -817.227039 0.0000000 (p < 0.01)

```{r}
library(ggpubr)
library(ggplot2)

ps.rich <- ggplot(
  ps.ps.alpha,
  aes(
    x = factor(Month_Year, levels = month_levels),
    y = Observed,
    fill = factor(Month_Year, levels = month_levels)
  )
) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "black") +
  scale_fill_manual(values = month_colors) +
  coord_cartesian(ylim = c(50, 3500)) +
  ylab("Richness") +
  xlab("Sampling Period") +
  theme_classic(base_size = 14) +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "none"
  ) +
  stat_compare_means(
    comparisons = list(c("1", "5"), c("3", "5")), 
    method = "t.test", 
    label = "p.signif", 
    label.y = c(7.3, 7.5)  # adjust y positions as needed
  )

ps.rich

# Save the plot
# ggsave("shannon.summer.alpha.png", plot = shannon.sum, width = 6, height = 5, dpi = 300)

```


Evenness
```{r}
library(vegan)
ps.alpha.4 <-aov(Evenness ~ Site * Month_Year, data = ps.ps.alpha)
summary(ps.alpha.4)

#Assumption checks
#1. Normality
qqnorm(residuals(ps.alpha.4))
qqline(residuals(ps.alpha.4), col = "red")
shapiro_test <- shapiro.test(residuals(ps.alpha.4))
print(shapiro_test)
#2. Homogeneity
library(car)
levene_test <- leveneTest(Evenness ~ Site * Month_Year, data = ps.ps.alpha)
print(levene_test)
#3 linearity
plot(ps.alpha.4, which = 1)
plot
#4 independence (think about sample design)
durbinWatsonTest(ps.alpha.4)
#The Durbin-Watson statistic ranges from 0 to 4.
#2 = no autocorrelation
#< 2 = positive autocorrelation
#> 2 = negative autocorrelation
```
*did not pass assumptions, so do KW

we have to do independent nonparametric kruskal wallis tests...... cannot look at the interaction between site and month! 
```{r}
kruskal.test(Evenness ~ Month_Year, data = ps.ps.alpha)  # main effect of month
kruskal.test(Evenness ~ Site, data = ps.ps.alpha)        # main effect of site
```
Yes to month_year!

checking for spreads or extreme skews...
```{r}
boxplot(Evenness ~ Month_Year, data = ps.ps.alpha)
boxplot(Evenness ~ Site, data = ps.ps.alpha)
#Shannon diversity distributions differ among months (rank-based differences)
```

looks fine to me! ~ let's do post hoc test for month_year
```{r}
library(FSA)
dunn_results_ps4 <- dunnTest(Evenness ~ Month_Year,
                         data = ps.ps.alpha,
                         method = "bh")   # adjust p-values
dunn_results_ps4
```
August_2024 - December_2023	-3.0668545	0.0021632410	0.015575335
August_2024 - July_2023	-2.8817149	0.0039551744	0.020340897	
August_2024 - June_2023	-3.5016187	0.0004624410	0.008323937	
April_2024 - July_2025	2.9113947	0.0035981913	0.021589148	
December_2023 - July_2025	3.4194114	0.0006275677	0.007530812
July_2023 - July_2025	3.2557357	0.0011309898	0.010178908
July_2025 - June_2023	-3.8164598	0.0001353801	0.004873685
July_2025 - October_2023	-2.5915263	0.0095551227	0.042998052	


```{r}
#month plot for even
ps.even <- ggplot(
  ps.ps.alpha,
  aes(
    x = factor(Month_Year, levels = month_levels),
    y = Evenness,
    fill = factor(Month_Year, levels = month_levels)
  )
) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "black") +
  scale_fill_manual(values = month_colors) +
  coord_cartesian(ylim = c(0.72, 1)) +
  ylab("Evenness") +
  xlab("Sampling Period") +
  theme_classic(base_size = 14) +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "none"
  ) +
  stat_compare_means(
    comparisons = list(c("1", "5"), c("3", "5")), 
    method = "t.test", 
    label = "p.signif", 
    label.y = c(7.3, 7.5)  # adjust y positions as needed
  )

ps.even

# Save the plot
# ggsave("shannon.summer.alpha.png", plot = shannon.sum, width = 6, height = 5, dpi = 300)

```
combo alpha plot:
```{r}
ps.alpha <- cowplot::plot_grid(
  ps.shannon,
  ps.rich,
  ps.even,
  ncol = 3,
  align = "h",
  rel_widths = c(1,1,1))  # optional panel labels

ps.alpha
```
```{r}
ggsave("ps_alpha_plot.pdf", plot = ps.alpha, width = 15, height = 8, dpi = 300)
```

how many samples in each month?
```{r}
table(ps.ps.alpha$Month_Year)
```


Nutrients 
TOM - stat results (sampling period & site) ~log transformed anova
TOC - No stat results
TN -  site result ~kruskal wallis test (nonparametric)
```{r}
ps.nuts.1 <- aov(TOM ~ Month_Year * Site, data = ps.ps.alpha)
summary(ps.nuts.1)
#assumption check
# 1. Normality
qqnorm(residuals(ps.nuts.1))
qqline(residuals(ps.nuts.1), col = "red")
shapiro.test(residuals(ps.nuts.1))

# 2. Homogeneity of variance
car::leveneTest(TOM ~ Site * Month_Year, data = ps.ps.alpha)
plot(ps.nuts.1, which = 3)

# 3. Residual diagnostics (model fit / variance patterns)
plot(ps.nuts.1, which = 1)
```
```{r}
ps.ps.alpha$TOM_log <- log1p(ps.ps.alpha$TOM)
ps.nuts.1log <- aov(TOM_log ~ Site * Month_Year, data = ps.ps.alpha)
summary(ps.nuts.1log)

# 1. Normality
qqnorm(residuals(ps.nuts.1log))
qqline(residuals(ps.nuts.1log), col = "red")
shapiro.test(residuals(ps.nuts.1log))

# 2. Homogeneity of variance
car::leveneTest(TOM_log ~ Site * Month_Year, data = ps.ps.alpha)
plot(ps.nuts.1log, which = 3)

# 3. Residual diagnostics (model fit / variance patterns)
plot(ps.nuts.1log, which = 1)
```
post hoc (month_year)
```{r}
TukeyHSD(ps.nuts.1log, which = "Site")
TukeyHSD(ps.nuts.1log, which = "Month_Year")
```
July_2023-June_2023          0.456737854  0.07950964  0.83396607 0.0078466
October_2023-July_2023      -0.411557988 -0.73824720 -0.08486877 0.0049045
December_2023-July_2023     -0.386175978 -0.71885999 -0.05349196 0.0124968
April_2024-July_2023        -0.415737761 -0.74242698 -0.08904854 0.0043294

July 2023 > June, October, Dec 2023 & April 2024

```{r}
#not using
ps.nuts.2 <- aov(TOC ~ Month_Year * Site, data = ps.ps.alpha)
summary(ps.nuts.2)
#assumption check
# 1. Normality
qqnorm(residuals(ps.nuts.2))
qqline(residuals(ps.nuts.2), col = "red")
shapiro.test(residuals(ps.nuts.2))

# 2. Homogeneity of variance
car::leveneTest(TOC ~ Site * Month_Year, data = ps.ps.alpha)
plot(ps.nuts.2, which = 3)

# 3. Residual diagnostics (model fit / variance patterns)
plot(ps.nuts.2, which = 1)
```
```{r}
#not using
ps.ps.alpha$TOC_log <- log1p(ps.ps.alpha$TOC)
ps.nuts.2log <- aov(TOC_log ~ Site * Month_Year, data = ps.ps.alpha)
summary(ps.nuts.2log)

# 1. Normality
qqnorm(residuals(ps.nuts.2log))
qqline(residuals(ps.nuts.2log), col = "red")
shapiro.test(residuals(ps.nuts.2log))

# 2. Homogeneity of variance
car::leveneTest(TOC_log ~ Site * Month_Year, data = ps.ps.alpha)
plot(ps.nuts.2log, which = 3)

# 3. Residual diagnostics (model fit / variance patterns)
plot(ps.nuts.2log, which = 1)
```

```{r}
#NONPARAMETRIC
kruskal.test(TOC ~ Month_Year, data = ps.ps.alpha)  # main effect of month
kruskal.test(TOC ~ Site, data = ps.ps.alpha)        # main effect of site
```
NONE 
checking for spreads or extreme skews...
```{r}
boxplot(TOC ~ Month_Year, data = ps.ps.alpha)
boxplot(TOC ~ Site, data = ps.ps.alpha)
```
looks fine to me! 

```{r}
ps.nuts.3 <- aov(TN ~ Month_Year * Site, data = ps.ps.alpha)
summary(ps.nuts.3)
#assumption check
# 1. Normality
qqnorm(residuals(ps.nuts.3))
qqline(residuals(ps.nuts.3), col = "red")
shapiro.test(residuals(ps.nuts.3))

# 2. Homogeneity of variance
car::leveneTest(TN ~ Site * Month_Year, data = ps.ps.alpha)
plot(ps.nuts.3, which = 3)

# 3. Residual diagnostics (model fit / variance patterns)
plot(ps.nuts.3, which = 1)
```
```{r}
#not using
ps.ps.alpha$TN_log <- log1p(ps.ps.alpha$TN)
ps.nuts.3log <- aov(TN_log ~ Site * Month_Year, data = ps.ps.alpha)
summary(ps.nuts.3log)

# 1. Normality
qqnorm(residuals(ps.nuts.3log))
qqline(residuals(ps.nuts.3log), col = "red")
shapiro.test(residuals(ps.nuts.3log))

# 2. Homogeneity of variance
car::leveneTest(TN_log ~ Site * Month_Year, data = ps.ps.alpha)
plot(ps.nuts.3log, which = 3)

# 3. Residual diagnostics (model fit / variance patterns)
plot(ps.nuts.3log, which = 1)
```

```{r}
#NONPARAMETRIC
kruskal.test(TN ~ Month_Year, data = ps.ps.alpha)  # main effect of month
kruskal.test(TN ~ Site, data = ps.ps.alpha)        # main effect of site
```
NONE 
checking for spreads or extreme skews...
```{r}
boxplot(TN ~ Month_Year, data = ps.ps.alpha)
boxplot(TN ~ Site, data = ps.ps.alpha)
```


Nutrient correlations & heatmap

~nutrient correlations 
to see families across the entire experiment:
```{r}
#note:iessmdffam ~ calculated and saved below ..need to  read back in here
psmdffam %>%
  group_by(Family) %>%
  summarise(Total_Abundance = sum(Abundance)) %>%
  arrange(desc(Total_Abundance))
```
These are the >2% most relative abundant families (percentage) 
```{r}
pstop_families <- psmdffam %>%
  group_by(Family) %>%
  summarise(Abundance = sum(Abundance)) %>%
  mutate(Percentage = (Abundance / sum(Abundance)) * 100) %>%
  filter(Percentage > 2) %>%
  arrange(desc(Percentage))

pstop_families
```
Calculating correlation coefficients based on Pearson
```{r}
ps.ps.alpha$Sample <- ps.ps.alpha$Sample.ID
```

```{r}
# Load libraries
library(dplyr)
library(tidyr)
library(ggplot2)

#ps.ps.alpha
#psmdffam

# Merge dataframes 
merged.ps <- merge(ps.ps.alpha, psmdffam, by = "Sample")

#Define top families (this is taxa w > 2% relative abundance) ---
specific_family_ps <- c("Rhodobacteraceae", "Flavobacteriaceae",
                     "Saprospiraceae","A4b","Cyclobacteriaceae",
                     "Puniceicoccaceae","Desulfocapsaceae", "Halieaceae", "Woeseiaceae", "Pirellulaceae")

# Filter for top families & remove NAs 
merged_clean_ps <- merged.ps %>%
  filter(Family %in% specific_family_ps) %>%
  drop_na(Abundance, TOM.x, TOC.x, TN.x)

#  Pivot nutrients to long format 
corr_df_long_ps <- merged_clean_ps %>%
  pivot_longer(
    cols = c(TOM.x, TOC.x, TN.x),
    names_to = "Variable",
    values_to = "Nutrient"
  )

# Calculate correlations + p-values per family × nutrient
corr_family_ps <- corr_df_long_ps %>%
  group_by(Family, Variable) %>%
  summarize(
    Correlation_Coefficient = ifelse(
      length(unique(Nutrient)) > 1,
      cor.test(Abundance, Nutrient, method = "pearson")$estimate,
      NA
    ),
    
    P_value = ifelse(
      length(unique(Nutrient)) > 1,
      cor.test(Abundance, Nutrient, method = "pearson")$p.value,
      NA
    ),
    
    .groups = "drop"
  )
#write.csv(corr_family_ps, "ps.corr_family.csv")

```
~sig stars added later~
corr_fam_ps uploaded to github for easy readability of sig correlations 


HEATMAP #decided not to add to final paper!
```{r}
corr_family_ps$Family <- factor(
  corr_family_ps$Family,
  levels = rev(sort(unique(corr_family_ps$Family)))
)
psheatmap_topfamily <- ggplot(corr_family_ps, aes(x = Variable, y = Family, fill = Correlation_Coefficient)) +
  geom_tile(color = "white") +
  
  scale_fill_gradientn(
  colors = c(
    "#440154",
    "#9E5DB8",
    "#35B779",
    "#187244"
  ),

  name = "Pearson's Correlation"
) +
  scale_y_discrete(position = "right") +
  labs(x = "Nutrients", y = "Bacterial Families") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    panel.border = element_blank(),
    legend.position = "right"
  )

psheatmap_topfamily
```


PAM

```{r}
pam <- read.csv("SM.PAM copy.csv")
```

```{r}
#deleting diel pam for this part 
ps.pam <- pam[-(89:123), ]
```


```{r}
ps.pam1 <- aov(Y.II. ~ Month._Year * Site, data = ps.pam)
summary(ps.pam1)
#assumption check
# 1. Normality
qqnorm(residuals(ps.pam1))
qqline(residuals(ps.pam1), col = "red")
shapiro.test(residuals(ps.pam1))

# 2. Homogeneity of variance
car::leveneTest(Y.II. ~ Site * Month._Year, data = ps.pam)
plot(ps.pam1, which = 3)

# 3. Residual diagnostics (model fit / variance patterns)
plot(ps.pam1, which = 1)
```
significant interaction between month_year*site

estimated marginal means to untangle interaction!
```{r}
library(emmeans)
#Does PSII yield differ among months within each site?
emm_month <- emmeans(ps.pam1, ~ Month._Year | Site)
pairs(emm_month, adjust = "tukey")
```
SB:
Aug_24 - Feb_24    -0.44542 0.0784 91  -5.685  <.0001
Feb_24 - July_23    0.38103 0.0679 91   5.615  <.0001
Feb_24 - June_24    0.33660 0.0784 91   4.296  0.0008
Feb_24 - Oct_23     0.34462 0.0898 91   3.839  0.0041

TM:
Dec_23 - July_23    0.27359 0.0798 91   3.426  0.0199

```{r}
#Does PSII yield differ among sites within each month?
emm_site <- emmeans(ps.pam1, ~ Site | Month._Year)
pairs(emm_site, adjust = "tukey")
```
Month._Year = Aug_24:
 contrast estimate     SE df t.ratio p.value
 SB - TM   -0.2008 0.0876 91  -2.292  0.0242
 
Month._Year = Feb_24:
 contrast estimate     SE df t.ratio p.value
 SB - TM    0.2085 0.0898 91   2.323  0.0224

**start back here!
pam plot:


~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~DS Stats!~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~


```{r}
di.pam <- read.csv("DI.PAM copy.csv")
```

